---
title: "WTI Options Portfolio: Full Greek P&L Attribution"
subtitle: "Individual Project"
date: "`r Sys.Date()`"
author: "Alex Hirtle (1719121)"
format:
  html:
    theme: cosmo
    embed-resources: true
    page-layout: full
    toc: true
    toc-location: left
    grid:
      margin-width: 200px
      body-width: 1000px
    resource:
      - "*.js"
      - "*.mp3"
editor: source
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
library(tidyverse)
library(gt)
library(RTL)
library(gridExtra)
library(patchwork)
library(plotly)

F_0  <- 73.00
r    <- 0.04
mult <- 1000

port <- dplyr::tribble(
  ~pos, ~type,   ~K,     ~T2M,   ~sigma,  ~n,
  "P1", "put",   70.00,  3/12,   0.3500,  +2000L,
  "P2", "put",   55.00,  3/12,   0.4200,  -2000L,
  "P3", "call",  80.00,  3/12,   0.3000,  -2000L
)
```

# Description

You are a risk analyst at an energy trading desk. The desk holds a portfolio of WTI crude oil options — specifically, options on NYMEX CL futures contracts.

Your role is to:

-   Build a full-Greek P&L attribution framework: the standard tool used on trading desks to explain daily P&L movements across **Delta** (directional exposure), **Gamma** (curvature), **Vega** (volatility sensitivity), **Vanna** (sensitivity of delta to volatility), and **Theta** (time decay).
-   You are expected to understand contracts and use the right option model and specifications in the `RTL` package.
-   Tell me in your own words the portfolio's risk profile, how the Greeks interact, and what this means for a desk managing this book in practice.
-   No AI use in any shape or form and assignment submitted should be a github repo that I can clone and assess.

# Objective

1.  Compute numerical Greeks for each position and for the overall portfolio across the full WTI price range (\$50–\$100/bbl).
2.  Interpret the portfolio's mark-to-market profile and explain how changing Gamma across strike levels affects delta-hedging in practice.
3.  Build a complete P&L attribution framework — Delta, Gamma, Vega, Vanna, and Theta — and verify it accurately explains overnight P&L under a realistic market move.
4.  Critically evaluate where and why the linear/quadratic approximation breaks down as spot moves away from current market.

# Exposure

```{r exposure_table, echo=FALSE}
dplyr::tribble(
  ~Position, ~Type,  ~`Strike K ($/bbl)`, ~Maturity, ~`Impl. Vol`, ~Contracts,
  "P1", "Put",  70.00, "3M", "35.00%", "+2,000 long",
  "P2", "Put",  55.00, "3M", "42.00%", "-2,000 short",
  "P3", "Call", 80.00, "3M", "30.00%", "-2,000 short"
) |>
  gt::gt() |>
  gt::tab_header(
    title    = "WTI Options Portfolio",
    subtitle = "Three-way collar: long put spread + short OTM call  |  F\u2080 = $73/bbl"
  ) |>
  gt::fmt_number(columns = `Strike K ($/bbl)`, decimals = 2)
```

**Market parameters**: current WTI futures price F₀ = **\$73.00/bbl**, risk-free rate r = **4.00%** (annualised, continuously compounded).

Select the appropriate pricing model from the `RTL` package for options on WTI crude oil futures. Consider the nature of the underlying contract — a NYMEX CL futures — when choosing the model and setting its cost-of-carry parameter.

# Question 1 (9 points)

For **each position**, compute the following Greeks **numerically** using central differences evaluated at `F_0`:

-   **Delta** (∂V/∂F)
-   **Gamma** (∂²V/∂F²)
-   **Vega** (∂V/∂σ)
-   **Theta** (∂V/∂T)

**Step size selection**: For each Greek, choose an appropriate finite-difference step size and explain your reasoning. A good step is small enough that the finite difference accurately approximates the true derivative, but not so small that floating-point subtraction errors dominate. Consider the natural scale of each risk factor (price in \$/bbl, vol as a decimal, time in years) and any special considerations for the short-dated options in this portfolio.

```{r, step sizes, echo = FALSE}

dplyr::tribble(
  ~ Greek, ~ `Step Size`,
  "Delta", "0.01",
  "Vega", "0.01",
  "Theta", "1/365"
) %>%
  gt::gt() 

```

Logic:

For delta I selected a step size of 0.01, as the minimum move in the underlying futures price is the same. I felt this step was small, but reasonable as selecting a larger size would lead to skipping possible values for the underlying.

For vega I chose a step size of 0.01, representing a 1% change in implied volatility of the contract. This step size felt small but not granular. Using a step size smaller than 0.01 would likely have caused more errors from floating point arithmetic, leading to a less accurate attribution later down the line.

For theta I decided on using a step size of 1 day. While that seems like quite a large jump, measuring theta with inter-day considerations of cahnge is not realistic or material. Using a smaller step size for theta would have likely increased the number of floating point errors introduced into the program.

**Dollar scaling**: Each raw numerical derivative gives the option price change **per barrel** for a unit move in the risk factor. To express the Greek as a total dollar P&L impact for a position, multiply by `n × mult`, where `n` is the signed contract count (positive = long, negative = short) and `mult = 1,000 bbl/contract`. For example, a delta of 0.50/bbl for a 3,000-contract long position means the desk gains or loses 0.50 × 3,000 × 1,000 = \$1,500,000 per \$1/bbl move in WTI.

Create plots showing each Greek vs. futures price over F ∈ \[50, 100\], with one line per position.

**d.** Compute the **portfolio mark-to-market (MTM)** value and **portfolio Gamma** across F ∈ \[50, 100\]. Plot them on two separate panels. In 3–5 sentences, explain:

-   What the MTM profile shape reveals about the portfolio's net directionality and curvature.
-   Where portfolio Gamma is highest and lowest across the price range, and why those regions correspond to specific positions.
-   What the changing Gamma means for a delta-hedger rebalancing daily as WTI moves through the \$50–\$100 range.

```{r q1}
# Step sizing
delta_step <- 0.01 # Minimum change in price on CL contract (As per CME CL contract specs)
vega_step <- 0.01 # 1% change in implied vol
theta_step <- 1/365 # 1 day (in years), 

# -------------------------------------------------------------------------------------------------------
position_greeks <- function(F_0, X, T2M, r, b = 0, sigma, step_F, step_S, step_T, type){
  # Gets the greeks at whatever the current Spot price is, given by F_0
  # Gotta run ts in another function to get portfolio greeks and shit
  # F_0: Spot
  # X: strike
  # T2M: time 2 maturity
  # r: rf rate
  # b: cost of carry
  # sigma: implied vol
  # step_F: step size for the spot
  # step_S: step size for vol
  # step_T: step size in time
  # type; call or put buster
  
  # Price at current spot
  current_price <- RTL::GBSOption(F_0, X, T2M, r, b, sigma, type)$price
  
  # Delta central diff
  price_up <- RTL::GBSOption(F_0 + step_F, X, T2M, r, b, sigma, type)$price
  price_down <- RTL::GBSOption(F_0 - step_F, X, T2M, r, b, sigma, type)$price
  
  delta <- (price_up - price_down) / (2 * step_F)

  # Gamma central diffe
  gamma <- (price_up - 2 * current_price + price_down) / (step_F ^ 2)
  
  # Vega - Vol up and down
  price_vol_up <- RTL::GBSOption(F_0, X, T2M, r, b, sigma + step_S, type)$price
  price_vol_down <- RTL::GBSOption(F_0, X, T2M, r, b, sigma - step_S, type)$price
  
  vega <- (price_vol_up - price_vol_down) / (2 * step_S)
  
  # Theta - Time forward :P
  price_t_up <- RTL::GBSOption(F_0, X, T2M - step_T, r, b, sigma, type)$price
  
  theta <- (price_t_up - current_price) / step_T
  
  # Vanna
  price_up_up <- RTL::GBSOption(F_0 + step_F, X, T2M, r, b, sigma + step_S, type)$price
  price_up_dn <- RTL::GBSOption(F_0 + step_F, X, T2M, r, b, sigma - step_S, type)$price
  price_dn_up <- RTL::GBSOption(F_0 - step_F, X, T2M, r, b, sigma + step_S, type)$price
  price_dn_dn <- RTL::GBSOption(F_0 - step_F, X, T2M, r, b, sigma - step_S, type)$price
  
  vanna <- (price_up_up - price_up_dn - price_dn_up + price_dn_dn) / (4 * step_F * step_S)
  
  
  
  price <- current_price # renaming this before putting it in the list, keep it pretty
  
  # Output list (KEEP THIS IN THE SAME ORDER OR THERE WILL BE CONSEQUENCES BRUH)
  greeks <- list(delta, gamma, vega, theta, vanna, price)
  
  return(greeks)
  
}

portfolio_greeks <- function(portfolio, r, b = 0, step_F, step_S, step_T, range = c(50, 100)) {
  # Gets the greeks for all the possible combinations of underlying spot price
  # Portfolio: port provided by phil
  # r: risk free rate
  # b: cost of carry
  # step_f: step size for futures price
  # step_s: step size for volatility
  # step_t: step size for time
  # range: range of values to test 
  
  # Number of contracts for each of the positions, keeps code below clearer
  P1_n <- portfolio$n[1]
  P2_n <- portfolio$n[2]
  P3_n <- portfolio$n[3]
  
  # Spine of the df, just has the spot value in range
  input_df <- data.frame(S = seq(range[1], range[2], by = step_F))
  
  # Grabs the list of all the greeks for the positions
  greeks_lists <- input_df %>%
    rowwise() %>%
    dplyr::mutate(P1_greeks = list(mapply(position_greeks, S, portfolio$K[1], portfolio$T2M[1], r, 0, portfolio$sigma[1], step_F, step_S, step_T, portfolio$type[1])),
                  P2_greeks = list(mapply(position_greeks, S, portfolio$K[2], portfolio$T2M[2], r, 0, portfolio$sigma[2], step_F, step_S, step_T, portfolio$type[2])),
                  P3_greeks = list(mapply(position_greeks, S, portfolio$K[3], portfolio$T2M[3], r, 0, portfolio$sigma[3], step_F, step_S, step_T, portfolio$type[3])))
  
  
  # This shit gets ugly here, im sorry phil, i couldnt think of any other way to do this part lol 
  # The columns with n are weighted to be total greek for each position,
  greeks_broken <- greeks_lists %>%
    dplyr::mutate(
      # Delta B-)
      P1_delta = P1_greeks[[1]],
      P1_delta_n = P1_delta * P1_n,
      P2_delta = P2_greeks[[1]],
      P2_delta_n = P2_delta * P2_n,
      P3_delta = P3_greeks[[1]],
      P3_delta_n = P3_delta * P3_n,
      
      # Gamma !!!!!
      P1_gamma = P1_greeks[[2]],
      P1_gamma_n = P1_gamma * P1_n,
      P2_gamma = P2_greeks[[2]],
      P2_gamma_n = P2_gamma * P2_n,
      P3_gamma = P3_greeks[[2]],
      P3_gamma_n = P3_gamma * P3_n,
      
      # Vega(s) O.O
      P1_vega = P1_greeks[[3]],
      P1_vega_n = P1_vega * P1_n,
      P2_vega = P2_greeks[[3]],
      P2_vega_n = P2_vega * P2_n,
      P3_vega = P3_greeks[[3]],
      P3_vega_n = P3_vega * P3_n,
      
      # Theta 
      P1_theta = P1_greeks[[4]],
      P1_theta_n = P1_theta * P1_n,
      P2_theta = P2_greeks[[4]],
      P2_theta_n = P2_theta * P2_n,
      P3_theta = P3_greeks[[4]],
      P3_theta_n = P3_theta * P3_n,
      
      # (Ha)Vanna
      P1_vanna = P1_greeks[[5]],
      P1_vanna_n = P1_vanna * P1_n,
      P2_vanna = P2_greeks[[5]],
      P2_vanna_n = P2_vanna * P2_n,
      P3_vanna = P3_greeks[[5]],
      P3_vanna_n = P3_vanna * P3_n,
      
      # Price
      P1_price = P1_greeks[[6]],
      P1_price_n = P1_price * P1_n,
      P2_price = P2_greeks[[6]],
      P2_price_n = P2_price * P2_n,
      P3_price = P3_greeks[[6]],
      P3_price_n = P3_price * P3_n,
      
      # Portfolio crap
      Port_delta_n = P1_delta_n + P2_delta_n + P3_delta_n,
      Port_gamma_n = P1_gamma_n + P2_gamma_n + P3_gamma_n,
      Port_vega_n = P1_vega_n + P2_vega_n + P3_vega_n,
      Port_theta_n = P1_theta_n + P2_theta_n + P3_theta_n,
      Port_vanna_n = P1_vanna_n + P2_vanna_n + P3_vanna_n,
      Port_price_n = P1_price_n + P2_price_n + P3_price_n
    ) %>%
    dplyr::select(-c(P1_greeks, P2_greeks, P3_greeks))
  
  return(greeks_broken)
  
}
# -------------------------------------------------------------------------------------------------------


# The main portfolio
portfolio <- portfolio_greeks(port, r, 0, delta_step, vega_step, theta_step)



# Create plots showing each Greek vs. futures price over F ∈ [50, 100], with one line per position.
delta_plot <- plotly::plot_ly(data = portfolio) %>%
  add_trace(
    x = ~ S,
    y = ~ P1_delta_n,
    type = "scatter",
    mode = "lines",
    name = "P1"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P2_delta_n,
    type = "scatter",
    mode = "lines",
    name = "P2"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P3_delta_n,
    type = "scatter",
    mode = "lines",
    name = "P3"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ Port_delta_n,
    type = "scatter",
    mode = "lines",
    name = "Portfolio"
  ) %>%
  layout(
    title = list(text = "Delta vs Futures Price."),
    xaxis = list(title = "Futures Price"),
    yaxis = list(title = "Delta, weighted")
  )

gamma_plot <- plotly::plot_ly(data = portfolio) %>%
  add_trace(
    x = ~ S,
    y = ~ P1_gamma_n,
    type = "scatter",
    mode = "lines",
    name = "P1"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P2_gamma_n,
    type = "scatter",
    mode = "lines",
    name = "P2"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P3_gamma_n,
    type = "scatter",
    mode = "lines",
    name = "P3"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ Port_gamma_n,
    type = "scatter",
    mode = "lines",
    name = "Portfolio"
  ) %>%
  layout(
    title = list(text = "Gamma vs Futures Price."),
    xaxis = list(title = "Futures Price"),
    yaxis = list(title = "Gamma, weighted")
  )

vega_plot <- plotly::plot_ly(data = portfolio) %>%
  add_trace(
    x = ~ S,
    y = ~ P1_vega_n,
    type = "scatter",
    mode = "lines",
    name = "P1"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P2_vega_n,
    type = "scatter",
    mode = "lines",
    name = "P2"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P3_vega_n,
    type = "scatter",
    mode = "lines",
    name = "P3"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ Port_vega_n,
    type = "scatter",
    mode = "lines",
    name = "Portfolio"
  ) %>%
  layout(
    title = list(text = "Vega vs Futures Price."),
    xaxis = list(title = "Futures Price"),
    yaxis = list(title = "Vega, weighted")
  )

theta_plot <- plotly::plot_ly(data = portfolio) %>%
  add_trace(
    x = ~ S,
    y = ~ P1_theta_n,
    type = "scatter",
    mode = "lines",
    name = "P1"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P2_theta_n,
    type = "scatter",
    mode = "lines",
    name = "P2"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P3_theta_n,
    type = "scatter",
    mode = "lines",
    name = "P3"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ Port_theta_n,
    type = "scatter",
    mode = "lines",
    name = "Portfolio"
  ) %>%
  layout(
    title = list(text = "Theta vs Futures Price."),
    xaxis = list(title = "Futures Price"),
    yaxis = list(title = "Theta, weighted")
  )

greeks_plot <- subplot(delta_plot, gamma_plot, 
                       vega_plot, theta_plot, 
                       nrows = 2, titleX = TRUE, 
                       titleY = TRUE, margin = 0.05) %>%
  layout(title = list(text = "Greeks vs Futures Price."),
         showlegend = FALSE)
# -----------------------------------------------------------------------------

# Mark to market !!!!! (part d)
pnl <- portfolio %>%
  dplyr::select(S, contains("Port")) %>%
  dplyr::mutate(PNL = Port_price_n * mult)

MTM_plot <- plotly::plot_ly(data = pnl) %>%
  add_trace(
    x = ~ S,
    y = ~ PNL,
    type = "scatter",
    mode = "lines",
    name = "Mark-to-market Value ($)"
  ) %>%
  layout(
    title = list(text = "Portfolio Mark-to-market value ($)"),
    xaxis = list(title = "Futures Price"),
    yaxis = list(title = "P&L")
  )

Port_gamma_plot <- plotly::plot_ly(data = pnl) %>%
  add_trace(
    x = ~ S,
    y = ~ Port_gamma_n * mult,
    type = "scatter",
    mode = "lines",
    name = "Portfolio Gamma"
  ) %>%
  layout(
    title = list(text = "Portfolio Gamma"),
    xaxis = list(title = "Futures Price"),
    yaxis = list(title = "Gamma P&L")
  )
# -----------------------------------------------------------------

max_gamma <- portfolio$S[portfolio$Port_gamma_n == max(portfolio$Port_gamma_n)]
min_gamma <- portfolio$S[portfolio$Port_gamma_n == min(portfolio$Port_gamma_n)]
```

```{r, q1 out MTM, echo = FALSE} 

MTM_plot

```
```{r, q1 out greeks, echo = FALSE}

greeks_plot

```
## Question 1 Answers:

Question: What \[does\] the MTM profile shape reveals about the portfolio's net directionality and curvature?

Answer: The Mark-to-market shape shows that the firm is extremely bearish on CL. Their combined position indicates a market view of a predicted drop in price, with very little margin for any rally in price. The curvature (delta, and gamma) of the portfolio indicate that the firm is net short on both delta and gamma. Their portfolio gamma over the range of \$50 - \$100 varies slightly, reaching a local maximum at \$64.97, and conversely is at a minimum at \$50. Additionally, because portfolio gamma has both changes in slope, as well as curvature, its clear that Speed would play a large part in unexplained PnL over the range. Portfolio delta has a similar story, being negative entirely throughout the range - reaching a local maximum at \$50, and a local minimum at \$100.

--------------------------------------------------------------------------------

Question: Where [is] portfolio Gamma [...] highest and lowest across the price range, and why those regions correspond to specific positions.

Answer: Portfolio gamma is minimized at the futures price of \$50, and maximized at a futures price of \$64.97. The minimum point of gamma is dominated by P2, the short put with strike price of \$55. This is mainly becuase gamma for both of the other positions is near-zero in comparison, leading to P2 making up the majority of the portfolio gamma at that point. Additionall, gamma for that particular position (P2) steadily decreases (increases for long position) throughout the range, reaching a minimum at the same point as portfolio gamma. The maximum for gamma is dominated by the firms position in P1, their long put with strike of \$70. P1 spikes sharply between futures price of ~\$57 and ~\$78, causing gamma within the portfolio to rise sharply as well.

--------------------------------------------------------------------------------

Question: What [does] the changing Gamma means for a delta-hedger rebalancing daily as WTI moves through the \$50–\$100 range?

Answer: For a delta-hedger, the sweeping changes in gamma exposure over the range of futures prices, even between \$73 +/- \$5 would be extremely detrimental. The portfolio would likely require multiple re-hedges a day, and the transaction costs alone for those hedging adjustments would likely eat a decent chunk of short-term profit. 

--------------------------------------------------------------------------------

# Question 2 (3 points)

Sum Greeks across positions to show **portfolio-level Greeks** at the current market. Present in a `gt::gt()` table. At approximately what futures price does the portfolio delta change sign?

```{r q2}
# Just where spot is current spot, makes shit easier for the tribble
q2 <- portfolio[portfolio$S == F_0,]

dplyr::tribble(
  ~ Position, ~ Delta, ~ Gamma, ~ Vega, ~ Theta, ~ Vanna, ~ Sum,
  "P1", q2$P1_delta_n, q2$P1_gamma_n, q2$P1_vega_n, q2$P1_theta_n, q2$P1_vanna_n, sum(q2$P1_delta_n, q2$P1_gamma_n, q2$P1_vega_n, q2$P1_theta_n, q2$P1_vanna_n),
  "P2", q2$P2_delta_n, q2$P2_gamma_n, q2$P2_vega_n, q2$P2_theta_n, q2$P2_vanna_n, sum(q2$P2_delta_n, q2$P2_gamma_n, q2$P2_vega_n, q2$P2_theta_n, q2$P2_vanna_n),
  "P3", q2$P3_delta_n, q2$P3_gamma_n, q2$P3_vega_n, q2$P3_theta_n, q2$P3_vanna_n, sum(q2$P3_delta_n, q2$P3_gamma_n, q2$P3_vega_n, q2$P3_theta_n, q2$P3_vanna_n),
  "Portfolio", q2$Port_delta_n, q2$Port_gamma_n, q2$Port_vega_n, q2$Port_theta_n, q2$Port_vanna_n, sum(q2$Port_delta_n, q2$Port_gamma_n, q2$Port_vega_n, q2$Port_theta_n, q2$Port_vanna_n)) %>%
  gt::gt() %>%
  gt::fmt_number(suffixing = TRUE, accounting = TRUE) %>%
  gt::tab_header(title = "Positional, and Portfolio Greeks.")
```

Question: At approximately what futures price does the portfolio delta change sign?

Answer: Looking at the chart, and after drawing out the payoff for the combined portfolio, it seems as though delta would never be positive. However, it approaches zero when the underlying futures price approaches zero. That being said, it does come incredibly close to zero at a value of approximately $17

--------------------------------------------------------------------------------

# Question 3 (8 points)

Overnight, US-Iran tensions escalate sharply. The US strikes Iranian nuclear facilities; Iran retaliates by mining the Strait of Hormuz. WTI gaps up at the open:

-   WTI futures: F moves from **73.00** to **90.00** (+\$17.00/bbl)
-   Implied volatility: each position's sigma spikes by **+0.20** (20 vol points)
-   Time: **1 calendar day** passes

For each position and for the portfolio total, compute:

1.  **Actual P&L**: full revaluation at new F, new sigma, new T. P&L = (new price − old price) × n × mult.
2.  **Attributed P&L**: ΔV ≈ (Delta × ΔF) + (½ × Gamma × ΔF²) + (Vega × Δσ) + (Vanna × ΔF × Δσ) + (Theta × Δt), scaled to dollars.
3.  **Unexplained residual**: Actual − Attributed.
4.  **Attribution breakdown**: what share (%) of Attributed P&L comes from each Greek?

**Vanna** (∂²V/∂F∂σ) measures how delta changes as volatility moves. With both a large spot jump and a large vol spike coinciding, Vanna is non-negligible. Use a central-difference cross-partial:

$$\text{Vanna} = \frac{V(F+h, \sigma+k) - V(F+h, \sigma-k) - V(F-h, \sigma+k) + V(F-h, \sigma-k)}{4 \cdot h \cdot k}$$

with h = `step_S` and k = `step_v`. The Vanna P&L contribution is Vanna × ΔF × Δσ.

Present all results in a `gt::gt()` table. Which position dominates the loss? Is the Greek attribution a reliable approximation here, and why or why not?

```{r q3}
# Changes
F_1 <- 90.00
delta_F <- F_1 - F_0
delta_sigma <- 0.2
delta_t <- 1/365

# !!! For each position and for the portfolio total, compute:
# Actual P&L: full revaluation at new F, new sigma, new T. P&L = (new price − old price) × n × mult.
# Attributed P&L: ΔV ≈ (Delta × ΔF) + (½ × Gamma × ΔF²) + (Vega × Δσ) + (Vanna × ΔF × Δσ) + (Theta × Δt), scaled to dollars.
# Unexplained residual: Actual − Attributed.
# Attribution breakdown: what share (%) of Attributed P&L comes from each Greek?

# Old price shit
old <- portfolio %>%
  dplyr::filter(S == F_0) %>%
  dplyr::pull(Port_price_n) * mult

# New price (requires new table)
q3_port <- port %>%
  dplyr::mutate(
    T2M = T2M - delta_t,
    sigma = sigma + delta_sigma)

q3_portfolio <- portfolio_greeks(q3_port, r, 0, delta_step, vega_step, theta_step) %>%
  dplyr::filter(S == F_1)

new <- q3_portfolio %>%
  dplyr::pull(Port_price_n) * mult

# Actual pnl 
Actual_pnl <- new - old

# Attributed pnl
q3_attributed <- portfolio %>%
  dplyr::filter(S == F_0) %>%
  dplyr::select(contains("_n")) # To grab all the adjustd greeks 

# Finds all the moves for all positions, gets the portfolio move, then gets attributed (expected) move
# again, im sorry this is ugly but i tried to write it like 4 different ways and it didnt work :(
Attributed_pnl <- q3_attributed %>%
  dplyr::select(-contains("price")) %>%
  dplyr::mutate(
                # Delta :T
                P1_delta_move = P1_delta_n * delta_F,
                P2_delta_move = P2_delta_n * delta_F,
                P3_delta_move = P3_delta_n * delta_F,
                
                # Gamma :DD
                P1_gamma_move = 0.5 * P1_gamma_n * delta_F^2,
                P2_gamma_move = 0.5 * P2_gamma_n * delta_F^2,
                P3_gamma_move = 0.5 * P3_gamma_n * delta_F^2,
                
                # Vega :$
                P1_vega_move = P1_vega_n * delta_sigma,
                P2_vega_move = P2_vega_n * delta_sigma,
                P3_vega_move = P3_vega_n * delta_sigma,
                
                # Vanna :M
                P1_vanna_move = P1_vanna_n * delta_F * delta_sigma,
                P2_vanna_move = P2_vanna_n * delta_F * delta_sigma,
                P3_vanna_move = P3_vanna_n * delta_F * delta_sigma,
                
                # Theta :/
                P1_theta_move = P1_theta_n * delta_t,
                P2_theta_move = P2_theta_n * delta_t,
                P3_theta_move = P3_theta_n * delta_t,
                
                # Totals :>
                P1_total_move = P1_delta_move + P1_gamma_move + P1_vega_move + P1_vanna_move + P1_theta_move,
                P2_total_move = P2_delta_move + P2_gamma_move + P2_vega_move + P2_vanna_move + P2_theta_move,
                P3_total_move = P3_delta_move + P3_gamma_move + P3_vega_move + P3_vanna_move + P3_theta_move,
                across(where(is.numeric), ~ .x * mult)) %>%
  dplyr::select(contains("move")) %>%
  dplyr::mutate(delta_move = P1_delta_move + P2_delta_move + P3_delta_move,
                gamma_move = P1_gamma_move + P2_gamma_move + P3_gamma_move,
                vega_move = P1_vega_move + P2_vega_move + P3_vega_move,
                vanna_move = P1_vanna_move + P2_vanna_move + P3_vanna_move,
                theta_move = P1_theta_move + P2_theta_move + P3_theta_move,
                total_move = delta_move + gamma_move + vega_move + vanna_move + theta_move)


Unexplained_pnl <- Actual_pnl - Attributed_pnl$total_move

# Gt table time !!!! :)

a <- Attributed_pnl # to make it more readable to call
dplyr::tribble(
  ~ Position, ~ Delta, ~ Gamma, ~ Vega, ~ Vanna, ~ Theta, ~ `% Attribution`,
  "P1 num.", a$P1_delta_move, a$P1_gamma_move, a$P1_vega_move, a$P1_vanna_move, a$P1_theta_move, a$P1_total_move / a$total_move * 100,
  "P2 num.", a$P2_delta_move, a$P2_gamma_move, a$P2_vega_move, a$P2_vanna_move, a$P2_theta_move, a$P2_total_move / a$total_move * 100,
  "P3 num.", a$P3_delta_move, a$P3_gamma_move, a$P3_vega_move, a$P3_vanna_move, a$P3_theta_move, a$P3_total_move / a$total_move * 100,
  "Portfolio (%)", a$delta_move / a$total_move * 100, a$gamma_move / a$total_move * 100, a$vega_move / a$total_move * 100, a$vanna_move / a$total_move * 100, a$theta_move / a$total_move * 100, 100L
) %>% gt::gt() %>%
  gt::fmt_number(suffixing = TRUE, accounting = TRUE) %>%
  gt::tab_header(title = "Portfolio Attribution Properties")
```

Question: Which position dominates the loss?

Answer: Through delta, P1 and P3 carry the attributions. Both reaching extremely negative numbers, causing PnL to drop sharply. However, generally in the portfolio P3, the firms short call dominates losses. From the table above, P3 leads >100% of the losses for the portfolio. With the other two positions clawing back some value. Heuristically, this makes sense. Because P3 is a short call, as price rises (especially sharply) the firm would be exposed to increasing losses. 

--------------------------------------------------------------------------------

Question: Is the Greek attribution a reliable approximation here, and why or why not?

Answer: Using greeks to approximate losses works reliably within a narrow range of underlying price fluctuations. As seen above, gamma exposure changes wildly throughout the price range, leading to Speed contributing a non-negligible amount to PnL. Because our approximation of the greeks was for a futures price approximately 19% lower than current prices, everything goes out the window. Greek PnL attribution estimation works well within small shifts of underlying value, but utterly fails in high market volatility.

--------------------------------------------------------------------------------

# Question 4 (4 points)

The Iran war scenario (ΔF = +\$17) is an extreme move. How well does the Greek approximation hold as WTI moves further from the current market?

Holding vol and time constant at their initial values, compute Actual P&L (full revaluation) and Attributed P&L (Delta + Gamma only) for the portfolio across ΔF ∈ \[–20, +20\] in \$1 increments.

**a.** Plot Actual, Attributed, and the Residual on the same chart. Mark the Iran war move (ΔF = +17) with a vertical line.

**b.** At what ΔF does the absolute residual first exceed 5% of the absolute Actual P&L? Report in a `gt::gt()` table. What does this tell you about using Greeks for risk management in extreme scenarios?

```{r q4}
# New step sizing
delta_step <- 1
range_step <- 20

# Full revaluation
revaluation_port <- portfolio_greeks(port, r, 0, delta_step, delta_sigma, delta_t, range = c(F_0 - range_step, F_0 + range_step))

revaluation_port_F_0 <- revaluation_port %>%
  dplyr::filter(S == F_0)

# Question A
old_port_val <- revaluation_port_F_0$Port_price_n * mult

# Actual Pnl
Actual_pnl_new <- revaluation_port %>%
  dplyr::select(S, Port_price_n) %>%
  dplyr::mutate(Port_pnl_act = Port_price_n * mult) %>%
  dplyr::select(S, Port_pnl_act) %>%
  dplyr::mutate(Port_pnl_act = Port_pnl_act - old_port_val)

# Attributed Pnl
delta_old <- revaluation_port_F_0$Port_delta_n
gamma_old <- revaluation_port_F_0$Port_gamma_n

Attributed_pnl_new <- data.frame(S = revaluation_port$S,
                                 delta = delta_old,
                                 gamma = gamma_old) %>%
  dplyr::mutate(delta_pnl = mult * delta * (S - F_0),
                gamma_pnl = mult * gamma * 0.5 * (S - F_0)^2,
                Port_pnl_att = delta_pnl + gamma_pnl) %>%
  dplyr::select(S, Port_pnl_att)

# Pnl crazytime df !!!!!
q4_pnl <- Actual_pnl_new %>%
  dplyr::left_join(., Attributed_pnl_new, join_by(S)) %>%
  dplyr::mutate(Unexplained_pnl = Port_pnl_act - Port_pnl_att)

# Plotly vertical line function
# Copied from https://plotly.com/r/horizontal-vertical-shapes/

vline <- function(x = 0, color = "green") {
  list(
    type = "line",
    y0 = 0,
    y1 = 1,
    yref = "paper",
    x0 = x,
    x1 = x,
    line = list(color = color, dash="line")
  )
}

# -----------------------------------------------------------

# Plot of this shiesse
q4_plot <- plotly::plot_ly(data = q4_pnl) %>%
  add_trace(
    x = ~ S,
    y = ~ Port_pnl_act,
    type = "scatter",
    mode = "lines",
    name = "Actual"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ Port_pnl_att,
    type = "scatter",
    mode = "lines",
    name = "Attributed"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ Unexplained_pnl,
    type = "scatter",
    mode = "lines",
    name = "Unexplained"
  ) %>%
  layout(
    title = list(text = "Pnl, Actual, Attributed, Unexplained"),
    xaxis = list(title = "Futures Price"),
    yaxis = list(title = "Pnl ($)"),
    shapes = list(vline(F_1, "red"))
  )

# Question B
# At what ΔF does the absolute residual first exceed 5% of the absolute Actual P&L

q4_b <- q4_pnl %>%
  dplyr::mutate(Unexplained_pct = abs(Unexplained_pnl) / abs(Port_pnl_act) * 100) %>%
  dplyr::filter(Unexplained_pct > 5) %>%
  dplyr::mutate(S = S - F_0)
```

**A.**

``` {r, q4 out pnl}

q4_plot

```

**B.**
```{r, q4 unexplained percentage out}

q4_b %>%
  dplyr::select(S, Unexplained_pct) %>%
  dplyr::rename("Δ Futures Price" = S, "Unexplained PnL (% of Attribution)" = Unexplained_pct) %>%
  gt::gt() %>%
  gt::fmt(columns = `Unexplained PnL (% of Attribution)`, fns = function(x){paste0(round(x, 2), "%")}) %>%
  gt::fmt(columns = `Δ Futures Price`, fns = function(x){paste0("$", x)})
```
Question: At what ΔF does the absolute residual first exceed 5% of the absolute Actual P&L?

Answer: As evidenced above, the first value where the unexplained percentage of returns is >5% when moving through the negative ΔF is \$60, or when ΔF = -\$13. Additionally, the value at which unexplained percentage of returns is >5% when moving through positive ΔF is $90, or when ΔF = \$17.

--------------------------------------------------------------------------------

Question: What does this tell you about using Greeks for risk management in extreme scenarios?

Answer: This increase in unexplained PnL steadily grows as the spot price drifts from the starting value. Critically, this steady increase in forecast error for PnL attribution can lead to improper hedging, and improper risk management. Large market value swings of assets are not uncommon, and relying on increasingly inaccurate metrics to assess exposure and risk reduce a portfolio managers ability to properly assess the landscape around them. Using Phil's examples of a dashboard in a car, if your greeks (gauges) get less accurate as you decrease or increase speed, you would be likely to end up in an accident.

--------------------------------------------------------------------------------

# Question 5 (6 points)

Discuss the key learnings and insights from a Greek risk management perspective. Your explanation must:

-   Interpret the **full revaluation MTM profile** from Q1(d): describe the portfolio's curvature structure across the \$50–\$100 range. Identify the regions of maximum and minimum Gamma. Explain what the changing Gamma means for a desk running this portfolio — specifically, how frequently delta rebalancing is needed as WTI moves through the strikes.
-   Interpret the **full revaluation P&L and Gamma profiles** from Q1(d): describe the portfolio's curvature structure and identify which positions drive positive vs negative Gamma regions.
-   Identify which position and which Greek drove the largest loss in the Iran war scenario, and explain why that was predictable from the portfolio structure.
-   Explain the role of **Vanna**: why is it non-trivial when both a large spot jump and a large vol spike occur simultaneously?
-   Explain why the Greek approximation (Q4) breaks down at extreme moves, and what this means for a real desk managing this book.

Pay particular attention to why textbook delta-hedging assumptions break down in geopolitical gap scenarios, and what risk tools a real energy desk would use instead.

*Your answer here.*

### MTM Profile, Analysis of Gamma

Throughout the range of futures prices for simulation the mark-to-market value of the portfolio varies wildly. Hitting a local maximum at \$50, and a minimum at \$100. The slope (delta) and convexity (gamma) are both negative throughout the entire range. This leads to a downward sloping, concave down looking graph. This MTM profile is a smoothed version of a simple payoff at maturity plot of the portfolio, indicating a bearish market-view on future oil prices.  

Gamma reaches a maximum at an underlying futures price of \$64.97, and a minimum at underlying futures price of \$50. The large swings and unpredictability of gamma would pose a large problem for traders on an option desk. Because the steep slope of gamma (speed) around the current spot price the overall portfolio would likely require multiple re-hedges a day to keep either delta, or gamma neutral. This requirement for frequent re-hedging introduces a large degree of friction to the position, eating up profits in transaction fees, and requiring more actions than other, less gamma sensitive portfolios. 

### Revaluation P&L, Gamma Profiles

The portfolios curvature along the range of spot prices is that of negative slope, and negative convexity. With delta remaining negative from \$50 to \$100, and gamma following a similar trend. At portfolio gammas maximum (least negative), the gamma level is driven primarily by P1. Because P1's gamma remains positive throughout the range, it has a strong influence over the shape of the overall portfolio gamma curve. When gamma is at its minimum (most negative), the value can be attributed nearly entirely to P2. Because P2 is a short call, once the spot price drops below its strike, the position starts to incur heavy losses. Overall, a large portion of the "up" swings in gamma can be chalked up to P1's violent gamma swings, and the "down" swings in gamma are either attributed to P3 when spot is high, and P2 when spot is below the strike price of P2.

### Iran war scenario

When the simulated shift in price and volatility happen, three greeks lead the decline in portfolio value. Delta, which accounted for 70% of the loss, gamma making up 11% of the movement, and vanna, leading 13% of the decline. These movements in value can be mostly nearly entirely attributed to P3, the firms short call position with a strike of \$80. This heuristically makes sense from an textbook standpoint, because as price increase, the payoff at maturity for a short call position declines linearly. Additionally, because the other two positions were puts with strikes well below the shifted price of \$90, the value of those positions become negligible in comparison to the short call. As the price of the underlying becomes more out of the money (OTM) for the puts, the value declines.  

### Vanna

Vanna represents a cross-sensitivity between both underlying price and volatility. On a portfolio level, vanna becomes increasingly more prominent within the PnL attribution framework as both of those values vary, particularly when they vary strongly together in the same direction. In the situation with the Iran war, both volatility and futures price rise sharply; leading to an extreme increase in the weight of vanna within the attribution.  

### Greek approximation

Using greeks to approximate portfolio moves works well with low volatility assets. However, when large price or volatility swings happen, the higher order greeks play a larger role in the attribution of PnL. A real trading desk would likely use a larger variety of greeks to more accurately simulate swings in value and volatility. Including speed, rho, and charm within a PnL attribution framework would increase validity of the model, as well as reduce error when estimating PnL changes through changing input variables. If a firm were to run this portfolio on their books, they'd need a comprehensive hedging and re-hedging strategy that both minimizes unwanted exposure to higher order greeks, as well as reduce overall transaction costs for the portfolio.

### Lessons Learned

Throughout the course of this project I learned a lot of new tricks, and enhanced my mental model of options pricing. Before I started this project, I didn't have a tangible grasp on exactly how all the greeks interacted, and how small changes in spot price can swing values so violently. 

I spent a lot of time going back over previously completed work as I talked to classmates about the requirements, and began to understand more of what I was measuring with a lot of the greeks and other metrics. I ended up entirely rewriting large sections of the project, completely scrapping what I had before and giving everything a fresh try with an improved understanding of my goals.

While I think I did a decent job of correctly implementing functions to measure greeks, I also feel I have more questions now about the math behind each of the greeks and their interactions and attributions. There are definitly some errors in logic, or incorrect calculations (particularly during the PnL attribution section) throughout my project.

I don't think I truly understood the depth and amount of time required to finish this project before I started. Going in, I felt like I had a firm grasp on all the greeks, and their interactions. However, shortly after starting, I realized I had a large gap in my knowledge between textbook understanding, and practical application. 

This project definitely sharpened my R skills, and opened my eyes to a wide world of solutions and mental models for completing the tasks given.

### Final Note

I really appreciate the time you put into making this project, and how everything is applicable to the world as it stands now. I find a lot of professors reuse old "toy" examples a lot for students that never allow them to make the connection between textbook and real life. However, with this project I think you really nailed it. It's interesting, difficult, and requires a great deal of learning beyond the notes we were given. It balances simplicity in assumptions with complexity in requirements very well, and I found the project fair, and very interesting.

Thanks Phil!
