---
title: "WTI Options Portfolio: Full Greek P&L Attribution"
subtitle: "Individual Project"
date: "`r Sys.Date()`"
author: "Alex Hirtle (1719121)"
format:
  html:
    theme: cosmo
    embed-resources: true
    page-layout: full
    toc: true
    toc-location: left
    grid:
      margin-width: 200px
      body-width: 1000px
    resource:
      - "*.js"
      - "*.mp3"
editor: source
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
library(tidyverse)
library(gt)
library(RTL)
library(gridExtra)
library(patchwork)

F_0  <- 73.00
r    <- 0.04
mult <- 1000

port <- dplyr::tribble(
  ~pos, ~type,   ~K,     ~T2M,   ~sigma,  ~n,
  "P1", "put",   70.00,  3/12,   0.3500,  +2000L,
  "P2", "put",   55.00,  3/12,   0.4200,  -2000L,
  "P3", "call",  80.00,  3/12,   0.3000,  -2000L
)
```

# Description

You are a risk analyst at an energy trading desk. The desk holds a portfolio of WTI crude oil options — specifically, options on NYMEX CL futures contracts.

Your role is to:

-   Build a full-Greek P&L attribution framework: the standard tool used on trading desks to explain daily P&L movements across **Delta** (directional exposure), **Gamma** (curvature), **Vega** (volatility sensitivity), **Vanna** (sensitivity of delta to volatility), and **Theta** (time decay).
-   You are expected to understand contracts and use the right option model and specifications in the `RTL` package.
-   Tell me in your own words the portfolio's risk profile, how the Greeks interact, and what this means for a desk managing this book in practice.
-   No AI use in any shape or form and assignment submitted should be a github repo that I can clone and assess.

# Objective

1.  Compute numerical Greeks for each position and for the overall portfolio across the full WTI price range (\$50–\$100/bbl).
2.  Interpret the portfolio's mark-to-market profile and explain how changing Gamma across strike levels affects delta-hedging in practice.
3.  Build a complete P&L attribution framework — Delta, Gamma, Vega, Vanna, and Theta — and verify it accurately explains overnight P&L under a realistic market move.
4.  Critically evaluate where and why the linear/quadratic approximation breaks down as spot moves away from current market.

# Exposure

```{r exposure_table, echo=FALSE}
dplyr::tribble(
  ~Position, ~Type,  ~`Strike K ($/bbl)`, ~Maturity, ~`Impl. Vol`, ~Contracts,
  "P1", "Put",  70.00, "3M", "35.00%", "+2,000 long",
  "P2", "Put",  55.00, "3M", "42.00%", "-2,000 short",
  "P3", "Call", 80.00, "3M", "30.00%", "-2,000 short"
) |>
  gt::gt() |>
  gt::tab_header(
    title    = "WTI Options Portfolio",
    subtitle = "Three-way collar: long put spread + short OTM call  |  F\u2080 = $73/bbl"
  ) |>
  gt::fmt_number(columns = `Strike K ($/bbl)`, decimals = 2)
```

**Market parameters**: current WTI futures price F₀ = **\$73.00/bbl**, risk-free rate r = **4.00%** (annualised, continuously compounded).

Select the appropriate pricing model from the `RTL` package for options on WTI crude oil futures. Consider the nature of the underlying contract — a NYMEX CL futures — when choosing the model and setting its cost-of-carry parameter.

# Question 1 (9 points)

For **each position**, compute the following Greeks **numerically** using central differences evaluated at `F_0`:

-   **Delta** (∂V/∂F)
-   **Gamma** (∂²V/∂F²)
-   **Vega** (∂V/∂σ)
-   **Theta** (∂V/∂T)

**Step size selection**: For each Greek, choose an appropriate finite-difference step size and explain your reasoning. A good step is small enough that the finite difference accurately approximates the true derivative, but not so small that floating-point subtraction errors dominate. Consider the natural scale of each risk factor (price in \$/bbl, vol as a decimal, time in years) and any special considerations for the short-dated options in this portfolio.

**Dollar scaling**: Each raw numerical derivative gives the option price change **per barrel** for a unit move in the risk factor. To express the Greek as a total dollar P&L impact for a position, multiply by `n × mult`, where `n` is the signed contract count (positive = long, negative = short) and `mult = 1,000 bbl/contract`. For example, a delta of 0.50/bbl for a 3,000-contract long position means the desk gains or loses 0.50 × 3,000 × 1,000 = \$1,500,000 per \$1/bbl move in WTI.

Create plots showing each Greek vs. futures price over F ∈ \[50, 100\], with one line per position.

**d.** Compute the **portfolio mark-to-market (MTM)** value and **portfolio Gamma** across F ∈ \[50, 100\]. Plot them on two separate panels. In 3–5 sentences, explain:

-   What the MTM profile shape reveals about the portfolio's net directionality and curvature.
-   Where portfolio Gamma is highest and lowest across the price range, and why those regions correspond to specific positions.
-   What the changing Gamma means for a delta-hedger rebalancing daily as WTI moves through the \$50–\$100 range.

```{r q1}
# Step sizing
delta_step <- 0.01 # Minimum change in price on CL contract (As per CME CL contract specs)
vega_step <- 0.01 # 1% change in implied vol
theta_step <- 1/365 # 1 day (in years), 

# Stationary Greeks (Central Difference Method)

get_cd_greeks <- function(X, T2M, r, b = 0, sigma, type, step, range = c(50, 100)) {
  # X: Strike price
  # T2M: Time to matrity
  # r: rf rate
  # b: cost of carry
  # sigma: standard deviataion
  # type: put or call, string
  # step: step size to use for the up and down part of the whole operation
  # range: from what F_0 to what F_0
  
  input_df <- data.frame(S = seq(range[1], range[2], by = step))
  
  output_df <- input_df %>%
    dplyr::mutate(option = purrr::map(S, ~ RTL::GBSOption(S = .x, X = X, T2M = T2M, r = r, b = b, sigma = sigma, type = type)),
                  delta = purrr::map_dbl(option, ~ .x$delta),
                  gamma = purrr::map_dbl(option, ~.x$gamma),
                  vega = purrr::map_dbl(option, ~.x$vega) / 100,
                  theta = purrr::map_dbl(option, ~.x$theta) / 365,
                  price = purrr::map_dbl(option, ~.x$price)) %>%
    dplyr::select(-option)
  
  return(output_df)
}
# -------------------------------------------------------------------------------------------------------
position_greeks <- function(F_0, X, T2M, r, b = 0, sigma, step_F, step_S, step_T, type){
  # Gets the greeks at whatever the current Spot price is, given by F_0
  # Gotta run ts in another function to get portfolio greeks and shit
  # F_0: Spot
  # X: strike
  # T2M: time 2 maturity
  # r: rf rate
  # b: cost of carry
  # sigma: implied vol
  # step_F: step size for the spot
  # step_S: step size for vol
  # step_T: step size in time
  # type; call or put buster
  
  # Price at current spot
  current_price <- RTL::GBSOption(F_0, X, T2M, r, b, sigma, type)$price
  
  # Delta central diff
  price_up <- RTL::GBSOption(F_0 + step_F, X, T2M, r, b, sigma, type)$price
  price_down <- RTL::GBSOption(F_0 - step_F, X, T2M, r, b, sigma, type)$price
  
  delta <- (price_up - price_down) / (2 * step_F)

  # Gamma central diffe
  gamma <- (price_up - 2 * current_price + price_down) / (step_F ^ 2)
  
  # Vega - Vol up and down
  price_vol_up <- RTL::GBSOption(F_0, X, T2M, r, b, sigma + step_S, type)$price
  price_vol_down <- RTL::GBSOption(F_0, X, T2M, r, b, sigma - step_S, type)$price
  
  vega <- (price_vol_up - price_vol_down) / (2 * step_S)
  
  # Theta - Time forward :P
  price_t_up <- RTL::GBSOption(F_0, X, T2M - step_T, r, b, sigma, type)$price
  
  theta <- (price_t_up - current_price) / step_T
  
  # Vanna
  price_up_up <- RTL::GBSOption(F_0 + step_F, X, T2M, r, b, sigma + step_S, type)$price
  price_up_dn <- RTL::GBSOption(F_0 + step_F, X, T2M, r, b, sigma - step_S, type)$price
  price_dn_up <- RTL::GBSOption(F_0 - step_F, X, T2M, r, b, sigma + step_S, type)$price
  price_dn_dn <- RTL::GBSOption(F_0 - step_F, X, T2M, r, b, sigma - step_S, type)$price
  
  vanna <- (price_up_up - price_up_dn - price_dn_up + price_dn_dn) / (4 * step_F * step_S)
  
  
  
  price <- current_price # renaming this before putting it in the list, keep it pretty
  
  # Output list (KEEP THIS IN THE SAME ORDER OR THERE WILL BE CONSEQUENCES BRUH)
  greeks <- list(delta, gamma, vega, theta, vanna, price)
  
  return(greeks)
  
}

portfolio_greeks <- function(portfolio, F_0, r, b = 0, step_F, step_S, step_T, range = c(50, 100)) {
  # Gets the greeks for all the possible combinations of underlying spot price
  
  # Number of contracts for each of the positions, keeps code below clearer
  P1_n <- portfolio$n[1]
  P2_n <- portfolio$n[2]
  P3_n <- portfolio$n[3]
  
  # Spine of the df, just has the spot value in range
  input_df <- data.frame(S = seq(range[1], range[2], by = step_F))
  
  # Grabs the list of all the greeks for the positions
  greeks_lists <- input_df %>%
    rowwise() %>%
    dplyr::mutate(P1_greeks = list(mapply(position_greeks, S, portfolio$K[1], portfolio$T2M[1], r, 0, portfolio$sigma[1], step_F, step_S, step_T, portfolio$type[1])),
                  P2_greeks = list(mapply(position_greeks, S, portfolio$K[2], portfolio$T2M[2], r, 0, portfolio$sigma[2], step_F, step_S, step_T, portfolio$type[2])),
                  P3_greeks = list(mapply(position_greeks, S, portfolio$K[3], portfolio$T2M[3], r, 0, portfolio$sigma[3], step_F, step_S, step_T, portfolio$type[3])))
  
  
  # This shit gets ugly here, im sorry phil, i couldnt think of any other way to do this part lol 
  # The columns with n are weighted to be total greek for each position,
  greeks_broken <- greeks_lists %>%
    dplyr::mutate(
      # Delta B-)
      P1_delta = P1_greeks[[1]],
      P1_delta_n = P1_delta * P1_n,
      P2_delta = P2_greeks[[1]],
      P2_delta_n = P2_delta * P2_n,
      P3_delta = P3_greeks[[1]],
      P3_delta_n = P3_delta * P3_n,
      
      # Gamma !!!!!
      P1_gamma = P1_greeks[[2]],
      P1_gamma_n = P1_gamma * P1_n,
      P2_gamma = P2_greeks[[2]],
      P2_gamma_n = P2_gamma * P2_n,
      P3_gamma = P3_greeks[[2]],
      P3_gamma_n = P3_gamma * P3_n,
      
      # Vega(s) O.O
      P1_vega = P1_greeks[[3]],
      P1_vega_n = P1_vega * P1_n,
      P2_vega = P2_greeks[[3]],
      P2_vega_n = P2_vega * P2_n,
      P3_vega = P3_greeks[[3]],
      P3_vega_n = P3_vega * P3_n,
      
      # Theta 
      P1_theta = P1_greeks[[4]],
      P1_theta_n = P1_theta * P1_n,
      P2_theta = P2_greeks[[4]],
      P2_theta_n = P2_theta * P2_n,
      P3_theta = P3_greeks[[4]],
      P3_theta_n = P3_theta * P3_n,
      
      # (Ha)Vanna
      P1_vanna = P1_greeks[[5]],
      P1_vanna_n = P1_vanna * P1_n,
      P2_vanna = P2_greeks[[5]],
      P2_vanna_n = P2_vanna * P2_n,
      P3_vanna = P3_greeks[[5]],
      P3_vanna_n = P3_vanna * P3_n,
      
      # Price
      P1_price = P1_greeks[[6]],
      P1_price_n = P1_price * P1_n,
      P2_price = P2_greeks[[6]],
      P2_price_n = P2_price * P1_n,
      P3_price = P3_greeks[[6]],
      P3_price_n = P3_price * P3_n,
      
      # Portfolio shit
      Port_delta_n = P1_delta_n + P2_delta_n + P3_delta_n,
      Port_gamma_n = P1_gamma_n + P2_gamma_n + P3_gamma_n,
      Port_vega_n = P1_vega_n + P2_vega_n + P3_vega_n,
      Port_theta_n = P1_theta_n + P2_theta_n + P3_theta_n,
      Port_vanna_n = P1_vanna_n + P2_vanna_n + P3_vanna_n,
      Port_price_n = P1_price_n + P2_price_n + P3_price_n
    ) %>%
    dplyr::select(-c(P1_greeks, P2_greeks, P3_greeks))
  
  return(greeks_broken)
  
}

portfolio <- portfolio_greeks(port, F_0, r, 0, delta_step, vega_step, theta_step)

# Create plots showing each Greek vs. futures price over F ∈ [50, 100], with one line per position.

delta_plot <- plotly::plot_ly(data = portfolio) %>%
  add_trace(
    x = ~ S,
    y = ~ P1_delta_n,
    type = "scatter",
    mode = "lines",
    name = "P1 Delta"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P2_delta_n,
    type = "scatter",
    mode = "lines",
    name = "P2 Delta"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P3_delta_n,
    type = "scatter",
    mode = "lines",
    name = "P3 Delta"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ Port_delta_n,
    type = "scatter",
    mode = "lines",
    name = "Portfolio Delta"
  ) %>%
  layout(
    title = list(text = "Delta vs Futures Price."),
    xaxis = list(title = "Futures Price"),
    yaxis = list(title = "Delta, weighted")
  )

gamma_plot <- plotly::plot_ly(data = portfolio) %>%
  add_trace(
    x = ~ S,
    y = ~ P1_gamma_n,
    type = "scatter",
    mode = "lines",
    name = "P1 Gamma"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P2_gamma_n,
    type = "scatter",
    mode = "lines",
    name = "P2 Gamma"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P3_gamma_n,
    type = "scatter",
    mode = "lines",
    name = "P3 Gamma"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ Port_gamma_n,
    type = "scatter",
    mode = "lines",
    name = "Portfolio Gamma"
  ) %>%
  layout(
    title = list(text = "Gamma vs Futures Price."),
    xaxis = list(title = "Futures Price"),
    yaxis = list(title = "Gamma, weighted")
  )

vega_plot <- plotly::plot_ly(data = portfolio) %>%
  add_trace(
    x = ~ S,
    y = ~ P1_vega_n,
    type = "scatter",
    mode = "lines",
    name = "P1 Vega"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P2_vega_n,
    type = "scatter",
    mode = "lines",
    name = "P2 Vega"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P3_vega_n,
    type = "scatter",
    mode = "lines",
    name = "P3 Vega"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ Port_vega_n,
    type = "scatter",
    mode = "lines",
    name = "Portfolio Vega"
  ) %>%
  layout(
    title = list(text = "Vega vs Futures Price."),
    xaxis = list(title = "Futures Price"),
    yaxis = list(title = "Vega, weighted")
  )

theta_plot <- plotly::plot_ly(data = portfolio) %>%
  add_trace(
    x = ~ S,
    y = ~ P1_theta_n,
    type = "scatter",
    mode = "lines",
    name = "P1 Theta"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P2_theta_n,
    type = "scatter",
    mode = "lines",
    name = "P2 Theta"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P3_theta_n,
    type = "scatter",
    mode = "lines",
    name = "P3 Theta"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ Port_theta_n,
    type = "scatter",
    mode = "lines",
    name = "Portfolio theta"
  ) %>%
  layout(
    title = list(text = "Theta vs Futures Price."),
    xaxis = list(title = "Futures Price"),
    yaxis = list(title = "Theta, weighted")
  )

delta_plot
gamma_plot
vega_plot
theta_plot

# Ouput table for the stationary greeks


```

# Question 2 (3 points)

Sum Greeks across positions to show **portfolio-level Greeks** at the current market. Present in a `gt::gt()` table. At approximately what futures price does the portfolio delta change sign?

```{r q2}

pnl %>%
  dplyr::filter(S == F_0) %>%
  dplyr::select(S, contains("P", ignore.case = FALSE), total_pnl) %>%
  gt::gt()

delta_zero <- pnl %>%
  dplyr::filter(P_delta == 0)
delta_zero
```

# Question 3 (8 points)

Overnight, US-Iran tensions escalate sharply. The US strikes Iranian nuclear facilities; Iran retaliates by mining the Strait of Hormuz. WTI gaps up at the open:

-   WTI futures: F moves from **73.00** to **90.00** (+\$17.00/bbl)
-   Implied volatility: each position's sigma spikes by **+0.20** (20 vol points)
-   Time: **1 calendar day** passes

For each position and for the portfolio total, compute:

1.  **Actual P&L**: full revaluation at new F, new sigma, new T. P&L = (new price − old price) × n × mult.
2.  **Attributed P&L**: ΔV ≈ (Delta × ΔF) + (½ × Gamma × ΔF²) + (Vega × Δσ) + (Vanna × ΔF × Δσ) + (Theta × Δt), scaled to dollars.
3.  **Unexplained residual**: Actual − Attributed.
4.  **Attribution breakdown**: what share (%) of Attributed P&L comes from each Greek?

**Vanna** (∂²V/∂F∂σ) measures how delta changes as volatility moves. With both a large spot jump and a large vol spike coinciding, Vanna is non-negligible. Use a central-difference cross-partial:

$$\text{Vanna} = \frac{V(F+h, \sigma+k) - V(F+h, \sigma-k) - V(F-h, \sigma+k) + V(F-h, \sigma-k)}{4 \cdot h \cdot k}$$

with h = `step_S` and k = `step_v`. The Vanna P&L contribution is Vanna × ΔF × Δσ.

Present all results in a `gt::gt()` table. Which position dominates the loss? Is the Greek attribution a reliable approximation here, and why or why not?

```{r q3}
# Changes
F_1 <- 90.00
delta_F <- F_1 - F_0
delta_sigma <- 0.2
delta_t <- 1/365

# 

```

```{r q3}
# Basically repeating everything above, just using different names and values for everything :)

# Some statics for the jump
F_1 <- 90.00
delta_F <- F_1 - F_0
vol_jump <- 0.2
delta_t <- 1/365

# Redoing shit after the jump, adding in shifts in everythign to the port df
port_jump <- port %>%
  dplyr::mutate(sigma = sigma + vol_jump,
                T2M = T2M - delta_t)

# Grabbing the greens again, but with the change. 
P1_greeks_jump <- get_cd_greeks(port_jump$K[1], port_jump$T2M[1], r, 0, port_jump$sigma[1], type = port_jump$type[1], step = delta_step, range = c(50, 100)) %>%
  dplyr::rename_with(~ paste0(.x, "_p1"), c(delta, gamma, vega, theta, price))
P2_greeks_jump <- get_cd_greeks(port_jump$K[2], port_jump$T2M[2], r, 0, port_jump$sigma[2], type = port_jump$type[2], step = delta_step, range = c(50, 100)) %>%
  dplyr::rename_with(~ paste0(.x, "_p2"), c(delta, gamma, vega, theta, price))
P3_greeks_jump <- get_cd_greeks(port_jump$K[3], port_jump$T2M[3], r, 0, port_jump$sigma[3], type = port_jump$type[3], step = delta_step, range = c(50, 100)) %>%
  dplyr::rename_with(~ paste0(.x, "_p3"), c(delta, gamma, vega, theta, price))

greeks__full_table_jump <- P1_greeks_jump %>%
  dplyr::left_join(., P2_greeks_jump, join_by(S)) %>%
  dplyr::left_join(., P3_greeks_jump, join_by(S)) %>%
  dplyr::select(S, contains("delta"), contains("gamma"), contains("vega"), contains("theta"), contains("rho"), contains("price"))

portfolio_greeks_jump <- greeks__full_table_jump %>%
  dplyr::mutate(P_delta = (port$n[1] * delta_p1) + (port$n[2] * delta_p2) + (port$n[3] * delta_p3),
                P_gamma = (port$n[1] * gamma_p1) + (port$n[2] * gamma_p2) + (port$n[3] * gamma_p3),
                P_vega = (port$n[1] * vega_p1) + (port$n[2] * vega_p2) + (port$n[3] * vega_p3),
                P_theta = (port$n[1] * theta_p1) + (port$n[2] * theta_p2) + (port$n[3] * theta_p3),
                P_Price = (port$n[1] * price_p1) + (port$n[2] * price_p2) + (port$n[3] * price_p3))

pnl_jump <- portfolio_greeks_jump %>%
  dplyr::mutate(delta_pnl = mult * P_delta,
                gamma_pnl = mult * P_gamma,
                vega_pnl = mult * P_vega,
                theta_pnl = mult * P_theta,
                total_pnl = mult * P_Price)


# Finding vanna of the ORIGINAL portfolio, at the original price. 
find_vanna <- function(F_0, delta_F, K, T2M, r, sigma, vol_jump, type) {
  
  up_up <- RTL::GBSOption(F_0 + delta_F, K, T2M, r, 0, sigma + vol_jump, type = type)$price
  up_dn <- RTL::GBSOption(F_0 + delta_F, K, T2M, r, 0, sigma - vol_jump, type = type)$price
  dn_up <- RTL::GBSOption(F_0 - delta_F, K, T2M, r, 0, sigma + vol_jump, type = type)$price
  dn_dn <- RTL::GBSOption(F_0 - delta_F, K, T2M, r, 0, sigma - vol_jump, type = type)$price
  
  vanna_out <- (up_up - up_dn - dn_up + dn_dn) / (4 * delta_F * vol_jump) # Same formula phil has in the doc
  return(vanna_out)
}
# P1

P1_va_curr <- find_vanna(F_0, delta_F, port$K[1], port$T2M[1], r, port$sigma[1], vol_jump, port$type[1])

# P2

P2_va_curr <- find_vanna(F_0, delta_F, port$K[2], port$T2M[2], r, port$sigma[2], vol_jump, port$type[2])

# P3

P3_va_curr <- find_vanna(F_0, delta_F, port$K[3], port$T2M[3], r, port$sigma[3], vol_jump, port$type[3])

# Portfolio

P_vanna <- (P1_va_curr * port$n[1]) + (P2_va_curr * port$n[2]) + (P3_va_curr * port$n[3])
# THIS VANNA IS ONLY FOR F_0 --- DONT USE IT ANYWHERE ELSE OR ITLL BE FUCKED UP


# -----------------------------------------------------------------------------------------------------------------------------------------

# Individual PNL change:
# Use Pnl_jump df, subtract value from Pnl df from that to find jump

# P1 - Prices pre and post shift, multiplied by exposure and number of contracts
P1_pre <- pnl %>%
  dplyr::filter(S == F_0) %>%
  dplyr::select(price_p1) %>%
  dplyr::pull() * mult * port$n[1]

P1_post <- pnl_jump %>%
  dplyr::filter(S == F_1) %>%
  dplyr::select(price_p1) %>%
  dplyr::pull() * mult * port$n[1]

P1_shift <- P1_post - P1_pre

# P2 - Prices pre and post shift, multiplied by exposure and number of contracts
P2_pre <- pnl %>%
  dplyr::filter(S == F_0) %>%
  dplyr::select(price_p2) %>%
  dplyr::pull() * mult * port$n[2]

P2_post <- pnl_jump %>%
  dplyr::filter(S == F_1) %>%
  dplyr::select(price_p2) %>%
  dplyr::pull() * mult * port$n[2]

P2_shift <- P2_post - P2_pre

# P3 - Prices pre and post shift, multiplied by exposure and number of contracts
P3_pre <- pnl %>%
  dplyr::filter(S == F_0) %>%
  dplyr::select(price_p3) %>%
  dplyr::pull() * mult * port$n[3]

P3_post <- pnl_jump %>%
  dplyr::filter(S == F_1) %>%
  dplyr::select(price_p3) %>%
  dplyr::pull() * mult * port$n[3]

P3_shift <- P3_post - P3_pre

Actual_pnl <- sum(P1_shift, P2_shift, P3_shift)
# ---------------------------------------------

# Attributed Pnl
# Using original PNL values from the pnl df

Attributed_pnl <- pnl %>%
  dplyr::filter(S == F_0) %>%
  dplyr::select(P_delta, P_gamma, P_vega, P_theta, total_pnl) %>%
  dplyr::mutate(Attributed_pnl = mult * ((P_delta * delta_F) + (P_gamma * 0.5 * delta_F^2) + (P_vega * vol_jump) + (P_vanna * delta_F * vol_jump) + (P_theta * delta_t))) %>%
  dplyr::pull(Attributed_pnl)

# Percentage breakdown for each greek

pnl_F_0 <- pnl %>%
  dplyr::filter(S == F_0) %>%
  dplyr::mutate(P_vanna = P_vanna)

delta_contribution <- mult * (pnl_F_0$P_delta * delta_F)
gamma_contribution <- mult * (pnl_F_0$P_gamma * 0.5 * delta_F^2)
vega_contribution <- mult * (pnl_F_0$P_vega * vol_jump)
vanna_contribution <- mult * (pnl_F_0$P_vanna * delta_F * vol_jump)
theta_contribution <- mult * (pnl_F_0$P_theta * delta_t)

delta_attributed_pnl <- delta_contribution / Attributed_pnl
gamma_attributed_pnl <- gamma_contribution / Attributed_pnl
vega_attributed_pnl <- vega_contribution / Attributed_pnl
vanna_attributed_pnl <- vanna_contribution / Attributed_pnl
theta_attributed_pnl <- theta_contribution / Attributed_pnl



# Unexplained Pnl
Unexplained_pnl <- Actual_pnl - Attributed_pnl
```

# Question 4 (4 points)

The Iran war scenario (ΔF = +\$17) is an extreme move. How well does the Greek approximation hold as WTI moves further from the current market?

Holding vol and time constant at their initial values, compute Actual P&L (full revaluation) and Attributed P&L (Delta + Gamma only) for the portfolio across ΔF ∈ \[–20, +20\] in \$1 increments.

**a.** Plot Actual, Attributed, and the Residual on the same chart. Mark the Iran war move (ΔF = +17) with a vertical line.

**b.** At what ΔF does the absolute residual first exceed 5% of the absolute Actual P&L? Report in a `gt::gt()` table. What does this tell you about using Greeks for risk management in extreme scenarios?

```{r q4}

```

# Question 5 (6 points)

Discuss the key learnings and insights from a Greek risk management perspective. Your explanation must:

-   Interpret the **full revaluation MTM profile** from Q1(d): describe the portfolio's curvature structure across the \$50–\$100 range. Identify the regions of maximum and minimum Gamma. Explain what the changing Gamma means for a desk running this portfolio — specifically, how frequently delta rebalancing is needed as WTI moves through the strikes.
-   Interpret the **full revaluation P&L and Gamma profiles** from Q1(d): describe the portfolio's curvature structure and identify which positions drive positive vs negative Gamma regions.
-   Identify which position and which Greek drove the largest loss in the Iran war scenario, and explain why that was predictable from the portfolio structure.
-   Explain the role of **Vanna**: why is it non-trivial when both a large spot jump and a large vol spike occur simultaneously?
-   Explain why the Greek approximation (Q4) breaks down at extreme moves, and what this means for a real desk managing this book.

Pay particular attention to why textbook delta-hedging assumptions break down in geopolitical gap scenarios, and what risk tools a real energy desk would use instead.

*Your answer here.*
