---
title: "WTI Options Portfolio: Full Greek P&L Attribution"
subtitle: "Individual Project"
date: "`r Sys.Date()`"
author: "Alex Hirtle (1719121)"
format:
  html:
    theme: cosmo
    embed-resources: true
    page-layout: full
    toc: true
    toc-location: left
    grid:
      margin-width: 200px
      body-width: 1000px
    resource:
      - "*.js"
      - "*.mp3"
editor: source
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
library(tidyverse)
library(gt)
library(RTL)
library(gridExtra)
library(patchwork)

F_0  <- 73.00
r    <- 0.04
mult <- 1000

port <- dplyr::tribble(
  ~pos, ~type,   ~K,     ~T2M,   ~sigma,  ~n,
  "P1", "put",   70.00,  3/12,   0.3500,  +2000L,
  "P2", "put",   55.00,  3/12,   0.4200,  -2000L,
  "P3", "call",  80.00,  3/12,   0.3000,  -2000L
)
```

# Description

You are a risk analyst at an energy trading desk. The desk holds a portfolio of WTI crude oil options — specifically, options on NYMEX CL futures contracts.

Your role is to:

-   Build a full-Greek P&L attribution framework: the standard tool used on trading desks to explain daily P&L movements across **Delta** (directional exposure), **Gamma** (curvature), **Vega** (volatility sensitivity), **Vanna** (sensitivity of delta to volatility), and **Theta** (time decay).
-   You are expected to understand contracts and use the right option model and specifications in the `RTL` package.
-   Tell me in your own words the portfolio's risk profile, how the Greeks interact, and what this means for a desk managing this book in practice.
-   No AI use in any shape or form and assignment submitted should be a github repo that I can clone and assess.

# Objective

1.  Compute numerical Greeks for each position and for the overall portfolio across the full WTI price range (\$50–\$100/bbl).
2.  Interpret the portfolio's mark-to-market profile and explain how changing Gamma across strike levels affects delta-hedging in practice.
3.  Build a complete P&L attribution framework — Delta, Gamma, Vega, Vanna, and Theta — and verify it accurately explains overnight P&L under a realistic market move.
4.  Critically evaluate where and why the linear/quadratic approximation breaks down as spot moves away from current market.

# Exposure

```{r exposure_table, echo=FALSE}
dplyr::tribble(
  ~Position, ~Type,  ~`Strike K ($/bbl)`, ~Maturity, ~`Impl. Vol`, ~Contracts,
  "P1", "Put",  70.00, "3M", "35.00%", "+2,000 long",
  "P2", "Put",  55.00, "3M", "42.00%", "-2,000 short",
  "P3", "Call", 80.00, "3M", "30.00%", "-2,000 short"
) |>
  gt::gt() |>
  gt::tab_header(
    title    = "WTI Options Portfolio",
    subtitle = "Three-way collar: long put spread + short OTM call  |  F\u2080 = $73/bbl"
  ) |>
  gt::fmt_number(columns = `Strike K ($/bbl)`, decimals = 2)
```

**Market parameters**: current WTI futures price F₀ = **\$73.00/bbl**, risk-free rate r = **4.00%** (annualised, continuously compounded).

Select the appropriate pricing model from the `RTL` package for options on WTI crude oil futures. Consider the nature of the underlying contract — a NYMEX CL futures — when choosing the model and setting its cost-of-carry parameter.

# Question 1 (9 points)

For **each position**, compute the following Greeks **numerically** using central differences evaluated at `F_0`:

-   **Delta** (∂V/∂F)
-   **Gamma** (∂²V/∂F²)
-   **Vega** (∂V/∂σ)
-   **Theta** (∂V/∂T)

**Step size selection**: For each Greek, choose an appropriate finite-difference step size and explain your reasoning. A good step is small enough that the finite difference accurately approximates the true derivative, but not so small that floating-point subtraction errors dominate. Consider the natural scale of each risk factor (price in \$/bbl, vol as a decimal, time in years) and any special considerations for the short-dated options in this portfolio.

**Dollar scaling**: Each raw numerical derivative gives the option price change **per barrel** for a unit move in the risk factor. To express the Greek as a total dollar P&L impact for a position, multiply by `n × mult`, where `n` is the signed contract count (positive = long, negative = short) and `mult = 1,000 bbl/contract`. For example, a delta of 0.50/bbl for a 3,000-contract long position means the desk gains or loses 0.50 × 3,000 × 1,000 = \$1,500,000 per \$1/bbl move in WTI.

Create plots showing each Greek vs. futures price over F ∈ \[50, 100\], with one line per position.

**d.** Compute the **portfolio mark-to-market (MTM)** value and **portfolio Gamma** across F ∈ \[50, 100\]. Plot them on two separate panels. In 3–5 sentences, explain:

-   What the MTM profile shape reveals about the portfolio's net directionality and curvature.
-   Where portfolio Gamma is highest and lowest across the price range, and why those regions correspond to specific positions.
-   What the changing Gamma means for a delta-hedger rebalancing daily as WTI moves through the \$50–\$100 range.

```{r q1}
# Step sizing
delta_step <- 0.01 # Minimum change in price on CL contract (As per CME CL contract specs)
gamma_step <- 0.01 # Minimum change in delta
vega_step <- 0.001
theta_step <- 1/365 # 1 day (in years), 

# Stationary Greeks (Central Difference Method)

get_cd_greeks <- function(X, T2M, r, b = 0, sigma, type, step, range = c(50, 100)) {
  # X: Strike price
  # T2M: Time to matrity
  # r: rf rate
  # b: cost of carry
  # sigma: standard deviataion
  # type: put or call, string
  # step: step size to use for the up and down part of the whole operation
  # range: from what F_0 to what F_0
  
  input_df <- data.frame(S = seq(range[1], range[2], by = step))
  
  output_df <- input_df %>%
    dplyr::mutate(option = purrr::map(S, ~ RTL::GBSOption(S = .x, X = X, T2M = T2M, r = r, b = b, sigma = sigma, type = type)),
                  delta = purrr::map_dbl(option, ~ .x$delta),
                  gamma = purrr::map_dbl(option, ~.x$gamma),
                  vega = purrr::map_dbl(option, ~.x$vega),
                  theta = purrr::map_dbl(option, ~.x$theta),
                  rho = purrr::map_dbl(option, ~.x$rho),
                  price = purrr::map_dbl(option, ~.x$price)) %>%
    dplyr::select(-option)
  
  return(output_df)
}

## Delta

P1_d_down <- RTL::GBSOption(F_0 - delta_step, port$K[1], port$T2M[1], r, 0, port$sigma[1], port$type[1])$price
P1_d_up <- RTL::GBSOption(F_0 + delta_step, port$K[1], port$T2M[1], r, 0, port$sigma[1], port$type[1])$price

P1_d_num <- (P1_d_up - P1_d_down) / (2 * delta_step) 
# --------------------------------------------------
P2_d_down <- RTL::GBSOption(F_0 - delta_step, port$K[2], port$T2M[2], r, 0, port$sigma[2], port$type[2])$price
P2_d_up <- RTL::GBSOption(F_0 + delta_step, port$K[2], port$T2M[2], r, 0, port$sigma[2], port$type[2])$price

P2_d_num <- (P2_d_up - P2_d_down) / (2 * delta_step)
# --------------------------------------------------
P3_d_down <- RTL::GBSOption(F_0 - delta_step, port$K[3], port$T2M[3], r, 0, port$sigma[3], port$type[3])$price
P3_d_up <- RTL::GBSOption(F_0 + delta_step, port$K[3], port$T2M[3], r, 0, port$sigma[3], port$type[3])$price
  
P3_d_num <- (P3_d_up - P3_d_down) / (2 * delta_step)

## Gamma

P1_g_down <- RTL::GBSOption(F_0 - gamma_step, port$K[1], port$T2M[1], r, 0, port$sigma[1], port$type[1])$price
P1_g_curr <- RTL::GBSOption(F_0, port$K[1], port$T2M[1], r, 0, port$sigma[1], port$type[1])$price
P1_g_up <- RTL::GBSOption(F_0 + gamma_step, port$K[1], port$T2M[1], r, 0, port$sigma[1], port$type[1])$price

P1_g_num <- (P1_g_up - 2 * P1_g_curr + P1_g_down) / (gamma_step ^ 2)
# ------------------------------------------------------------------

P2_g_down <- RTL::GBSOption(F_0 - gamma_step, port$K[2], port$T2M[2], r, 0, port$sigma[2], port$type[2])$price
P2_g_curr <- RTL::GBSOption(F_0, port$K[2], port$T2M[2], r, 0, port$sigma[2], port$type[2])$price
P2_g_up <- RTL::GBSOption(F_0 + gamma_step, port$K[2], port$T2M[2], r, 0, port$sigma[2], port$type[2])$price

P2_g_num <- (P2_g_up - 2 * P2_g_curr + P2_g_down) / (gamma_step ^ 2)
# ------------------------------------------------------------------

P3_g_down <- RTL::GBSOption(F_0 - gamma_step, port$K[3], port$T2M[3], r, 0, port$sigma[3], port$type[3])$price
P3_g_curr <- RTL::GBSOption(F_0, port$K[3], port$T2M[3], r, 0, port$sigma[3], port$type[3])$price
P3_g_up <- RTL::GBSOption(F_0 + gamma_step, port$K[3], port$T2M[3], r, 0, port$sigma[3], port$type[3])$price

P3_g_num <- (P3_g_up - 2 * P3_g_curr + P3_g_down) / (gamma_step ^ 2)

## Vega gt::gt() %>%
  gt::tab_header(title = "WTI Option basket - Stationary Greeks")

P1_v_down <- RTL::GBSOption(F_0, port$K[1], port$T2M[1], r, 0, port$sigma[1] - vega_step, port$type[1])$price
P1_v_up <- RTL::GBSOption(F_0, port$K[1], port$T2M[1], r, 0, port$sigma[1] + vega_step, port$type[1])$price

P1_v_num <- (P1_v_up - P1_v_down) / (2 * vega_step * 100)
# -------------------------------------------------

P2_v_down <- RTL::GBSOption(F_0, port$K[2], port$T2M[2], r, 0, port$sigma[2] - vega_step, port$type[2])$price
P2_v_up <- RTL::GBSOption(F_0, port$K[2], port$T2M[2], r, 0, port$sigma[2] + vega_step, port$type[2])$price

P2_v_num <- (P2_v_up - P2_v_down) / (2 * vega_step * 100)
# -------------------------------------------------

P3_v_down <- RTL::GBSOption(F_0, port$K[3], port$T2M[2], r, 0, port$sigma[3] - vega_step, port$type[3])$price
P3_v_up <- RTL::GBSOption(F_0, port$K[3], port$T2M[3], r, 0, port$sigma[3] + vega_step, port$type[3])$price

P3_v_num <- (P3_v_up - P3_v_down) / (2 * vega_step * 100)

## Theta

P1_t_up <- RTL::GBSOption(F_0, port$K[1], port$T2M[1] - theta_step, r, 0, port$sigma[1], port$type[1])$price
P1_t_curr <- RTL::GBSOption(F_0, port$K[1], port$T2M[1], r, 0, port$sigma[1], port$type[1])$price

P1_t_num <- P1_t_up - P1_t_curr
# -----------------------------

P2_t_up <- RTL::GBSOption(F_0, port$K[2], port$T2M[2] - theta_step, r, 0, port$sigma[2], port$type[2])$price
P2_t_curr <- RTL::GBSOption(F_0, port$K[2], port$T2M[2], r, 0, port$sigma[2], port$type[2])$price

P2_t_num <- P2_t_up - P2_t_curr
# -----------------------------

P3_t_up <- RTL::GBSOption(F_0, port$K[3], port$T2M[3] - theta_step, r, 0, port$sigma[3], port$type[3])$price
P3_t_curr <- RTL::GBSOption(F_0, port$K[3], port$T2M[3], r, 0, port$sigma[3], port$type[3])$price

P3_t_num <- P3_t_up - P3_t_curr
# -----------------------------

# Ouput table for the stationary greeks

a <- dplyr::tribble(
  ~Position, ~Type, ~Delta, ~Gamma, ~Vega, ~Theta,
  "P1", port$type[1], P1_d_num, P1_g_num, P1_v_num, P1_t_num,
  "P2", port$type[2], P2_d_num, P2_g_num, P2_v_num, P2_t_num,
  "P3", port$type[3], P3_d_num, P3_g_num, P3_v_num, P3_t_num,
) %>%
  gt::gt() %>%
  gt::tab_header(title = "WTI Option basket - Stationary Greeks")


# Graphing the greeks

# gets the greeks for each position
P1_greeks <- get_cd_greeks(port$K[1], port$T2M[1], r, 0, port$sigma[1], type = port$type[1], step = delta_step, range = c(50, 100)) %>%
  dplyr::rename_with(~ paste0(.x, "_p1"), c(delta, gamma, vega, theta, rho, price))
P2_greeks <- get_cd_greeks(port$K[2], port$T2M[2], r, 0, port$sigma[2], type = port$type[2], step = delta_step, range = c(50, 100)) %>%
  dplyr::rename_with(~ paste0(.x, "_p2"), c(delta, gamma, vega, theta, rho, price))
P3_greeks <- get_cd_greeks(port$K[3], port$T2M[3], r, 0, port$sigma[3], type = port$type[3], step = delta_step, range = c(50, 100)) %>%
  dplyr::rename_with(~ paste0(.x, "_p3"), c(delta, gamma, vega, theta, rho, price))

# chucks them all in 1 df and sorts it so its more readable
greeks_full_table <- P1_greeks %>%
  dplyr::left_join(., P2_greeks, join_by(S)) %>%
  dplyr::left_join(., P3_greeks, join_by(S)) %>%
  dplyr::select(S, contains("delta"), contains("gamma"), contains("vega"), contains("theta"), contains("rho"), contains("price"))

delta_plots <- plotly::plot_ly(data = greeks_full_table) %>%
  add_trace(x = ~ S,
            y = ~ delta_p1,
            type = "scatter",
            mode = "lines",
            name = "P1 Delta") %>%
  add_trace(x = ~ S,
            y = ~ delta_p2,
            type = "scatter",
            mode = "lines",
            name = "P2 Delta") %>%
  add_trace(x = ~ S,
            y = ~ delta_p3,
            type = "scatter",
            mode = "lines",
            name = "P3 Delta") %>%
  layout(
    title = list(text = paste0("Position Delta - Step size ", {delta_step})),
    xaxis = list(title = paste0("Futures Price, S")),
    yaxis = list(title = paste0("Delta"))
  )

gamma_plots <- plotly::plot_ly(data = greeks_full_table) %>%
  add_trace(x = ~ S,
            y = ~ gamma_p1,
            type = "scatter",
            mode = "lines",
            name = "P1 Gamma") %>%
  add_trace(x = ~ S,
            y = ~ gamma_p2,
            type = "scatter",
            mode = "lines",
            name = "P2 Gamma") %>%
  add_trace(x = ~ S,
            y = ~ gamma_p3,
            type = "scatter",
            mode = "lines",
            name = "P3 Gamma") %>%
  layout(
    title = list(text = paste0("Position Gamma - Step size ", {gamma_step})),
    xaxis = list(title = paste0("Futures Price, S")),
    yaxis = list(title = paste0("Gamma")))

vega_plots <- plotly::plot_ly(data = greeks_full_table) %>%
  add_trace(x = ~ S,
            y = ~ vega_p1,
            type = "scatter",
            mode = "lines",
            name = "P1 Vega") %>%
  add_trace(x = ~ S,
            y = ~ vega_p2,
            type = "scatter",
            mode = "lines",
            name = "P2 Vega") %>%
  add_trace(x = ~ S,
            y = ~ vega_p3,
            type = "scatter",
            mode = "lines",
            name = "P3 Vega") %>%
  layout(
    title = list(text = paste0("Position Vega - Step size ", {vega_step})),
    xaxis = list(title = paste0("Futures Price, S")),
    yaxis = list(title = paste0("Vega")))

theta_plots <- plotly::plot_ly(data = greeks_full_table) %>%
  add_trace(x = ~ S,
            y = ~ theta_p1,
            type = "scatter",
            mode = "lines",
            name = "P1 Theta") %>%
  add_trace(x = ~ S,
            y = ~ theta_p2,
            type = "scatter",
            mode = "lines",
            name = "P2 Theta") %>%
  add_trace(x = ~ S,
            y = ~ theta_p3,
            type = "scatter",
            mode = "lines",
            name = "P3 Theta") %>%
  layout(
    title = list(text = paste0("Position Theta - Step size 1 day")),
    xaxis = list(title = paste0("Futures Price, S")),
    yaxis = list(title = paste0("Theta")))

# This the master plot :D
greek_plots <- subplot(
  delta_plots, gamma_plots,
  vega_plots, theta_plots,
  nrows = 2
)
greek_plots
# ---------------------------------------------------------

# Portfolio MTM

# Entire portfolio greeks - Didn't weight because they're all 2,000 contracts in a direction so fuck it i guess
portfolio_greeks <- greeks_full_table %>%
  dplyr::mutate(P_delta = (port$n[1] * delta_p1) + (port$n[2] * delta_p2) + (port$n[3] * delta_p3),
                P_gamma = (port$n[1]^2 * gamma_p1 * 0.5) + (port$n[2]^2 * gamma_p2 * 0.5) + (port$n[3]^2 * gamma_p3 * 0.5),
                P_vega = (port$n[1] * vega_p1) + (port$n[2] * vega_p2) + (port$n[3] * vega_p3),
                P_theta = (port$n[1] * theta_p1) + (port$n[2] * theta_p2) + (port$n[3] * theta_p3),
                P_Price = (port$n[1] * price_p1) + (port$n[2] * price_p2) + (port$n[3] * price_p3)) %>%
  dplyr::select(S, contains("P", ignore.case = FALSE)) # Cheeky select using ignore.case = False for the first time in my life

pnl <- portfolio_greeks %>%
  dplyr::mutate(delta_pnl = mult * P_delta,
                gamma_pnl = mult * P_gamma,
                vega_pnl = mult * P_vega,
                theta_pnl = mult * P_theta,
                total_pnl = mult * P_Price)

# Portfolio Gamma, Highs and lows

portfolio_gamma_plot <- plotly::plot_ly(data = pnl) %>%
  add_trace(x = ~ S,
            y = ~ P_gamma, 
            type = "scatter",
            mode = "lines",
            name = "Portfolio Gamma") %>%
  layout(title = list(text = "Portfolio Gamma"),
         xaxis = list(title = "Futures price, S"),
         yaxis = list(title = "Gamma"))

# Max and min portfolio gamma, and thier corresponding S
max_port_gamma <- pnl %>%
  dplyr::filter(P_gamma == max(P_gamma)) %>%
  dplyr::select(S, P_gamma)

min_port_gamma <- pnl %>%
  dplyr::filter(P_gamma == min(P_gamma)) %>%
  dplyr::select(S, P_gamma)
# ------------------------------------------------------

# Making a little table for our max and min gamma values
dplyr::tribble(~Side, ~S, ~Gamma,
               "Minimum", min_port_gamma[1], min_port_gamma[2],
               "Maximum", max_port_gamma[1], max_port_gamma[2]) %>%
  gt::gt() %>%
  gt::tab_header(title = "Max and min gamma, across S of $50 - $100")
# ------------------------------------------------------

pnl_plot <- plotly::plot_ly(data = pnl) %>%
  add_trace(
    x = ~ S,
    y = ~ total_pnl,
    type = "scatter",
    mode = "lines",
    name = "PNL"
  )
```

# Question 2 (3 points)

Sum Greeks across positions to show **portfolio-level Greeks** at the current market. Present in a `gt::gt()` table. At approximately what futures price does the portfolio delta change sign?

```{r q2}

pnl %>%
  dplyr::filter(S == F_0) %>%
  dplyr::select(S, contains("P", ignore.case = FALSE), total_pnl)

delta_zero <- pnl %>%
  dplyr::filter(P_delta >= 0)
delta_zero
```

# Question 3 (8 points)

Overnight, US-Iran tensions escalate sharply. The US strikes Iranian nuclear facilities; Iran retaliates by mining the Strait of Hormuz. WTI gaps up at the open:

-   WTI futures: F moves from **73.00** to **90.00** (+\$17.00/bbl)
-   Implied volatility: each position's sigma spikes by **+0.20** (20 vol points)
-   Time: **1 calendar day** passes

For each position and for the portfolio total, compute:

1.  **Actual P&L**: full revaluation at new F, new sigma, new T. P&L = (new price − old price) × n × mult.
2.  **Attributed P&L**: ΔV ≈ (Delta × ΔF) + (½ × Gamma × ΔF²) + (Vega × Δσ) + (Vanna × ΔF × Δσ) + (Theta × Δt), scaled to dollars.
3.  **Unexplained residual**: Actual − Attributed.
4.  **Attribution breakdown**: what share (%) of Attributed P&L comes from each Greek?

**Vanna** (∂²V/∂F∂σ) measures how delta changes as volatility moves. With both a large spot jump and a large vol spike coinciding, Vanna is non-negligible. Use a central-difference cross-partial:

$$\text{Vanna} = \frac{V(F+h, \sigma+k) - V(F+h, \sigma-k) - V(F-h, \sigma+k) + V(F-h, \sigma-k)}{4 \cdot h \cdot k}$$

with h = `step_S` and k = `step_v`. The Vanna P&L contribution is Vanna × ΔF × Δσ.

Present all results in a `gt::gt()` table. Which position dominates the loss? Is the Greek attribution a reliable approximation here, and why or why not?

```{r q3}



```

# Question 4 (4 points)

The Iran war scenario (ΔF = +\$17) is an extreme move. How well does the Greek approximation hold as WTI moves further from the current market?

Holding vol and time constant at their initial values, compute Actual P&L (full revaluation) and Attributed P&L (Delta + Gamma only) for the portfolio across ΔF ∈ \[–20, +20\] in \$1 increments.

**a.** Plot Actual, Attributed, and the Residual on the same chart. Mark the Iran war move (ΔF = +17) with a vertical line.

**b.** At what ΔF does the absolute residual first exceed 5% of the absolute Actual P&L? Report in a `gt::gt()` table. What does this tell you about using Greeks for risk management in extreme scenarios?

```{r q4}

```

# Question 5 (6 points)

Discuss the key learnings and insights from a Greek risk management perspective. Your explanation must:

-   Interpret the **full revaluation MTM profile** from Q1(d): describe the portfolio's curvature structure across the \$50–\$100 range. Identify the regions of maximum and minimum Gamma. Explain what the changing Gamma means for a desk running this portfolio — specifically, how frequently delta rebalancing is needed as WTI moves through the strikes.
-   Interpret the **full revaluation P&L and Gamma profiles** from Q1(d): describe the portfolio's curvature structure and identify which positions drive positive vs negative Gamma regions.
-   Identify which position and which Greek drove the largest loss in the Iran war scenario, and explain why that was predictable from the portfolio structure.
-   Explain the role of **Vanna**: why is it non-trivial when both a large spot jump and a large vol spike occur simultaneously?
-   Explain why the Greek approximation (Q4) breaks down at extreme moves, and what this means for a real desk managing this book.

Pay particular attention to why textbook delta-hedging assumptions break down in geopolitical gap scenarios, and what risk tools a real energy desk would use instead.

*Your answer here.*
