---
title: "WTI Options Portfolio: Full Greek P&L Attribution"
subtitle: "Individual Project"
date: "`r Sys.Date()`"
author: "Alex Hirtle (1719121)"
format:
  html:
    theme: cosmo
    embed-resources: true
    page-layout: full
    toc: true
    toc-location: left
    grid:
      margin-width: 200px
      body-width: 1000px
    resource:
      - "*.js"
      - "*.mp3"
editor: source
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
library(tidyverse)
library(gt)
library(RTL)
library(gridExtra)
library(patchwork)

F_0  <- 73.00
r    <- 0.04
mult <- 1000

port <- dplyr::tribble(
  ~pos, ~type,   ~K,     ~T2M,   ~sigma,  ~n,
  "P1", "put",   70.00,  3/12,   0.3500,  +2000L,
  "P2", "put",   55.00,  3/12,   0.4200,  -2000L,
  "P3", "call",  80.00,  3/12,   0.3000,  -2000L
)
```

# Description

You are a risk analyst at an energy trading desk. The desk holds a portfolio of WTI crude oil options — specifically, options on NYMEX CL futures contracts.

Your role is to:

-   Build a full-Greek P&L attribution framework: the standard tool used on trading desks to explain daily P&L movements across **Delta** (directional exposure), **Gamma** (curvature), **Vega** (volatility sensitivity), **Vanna** (sensitivity of delta to volatility), and **Theta** (time decay).
-   You are expected to understand contracts and use the right option model and specifications in the `RTL` package.
-   Tell me in your own words the portfolio's risk profile, how the Greeks interact, and what this means for a desk managing this book in practice.
-   No AI use in any shape or form and assignment submitted should be a github repo that I can clone and assess.

# Objective

1.  Compute numerical Greeks for each position and for the overall portfolio across the full WTI price range (\$50–\$100/bbl).
2.  Interpret the portfolio's mark-to-market profile and explain how changing Gamma across strike levels affects delta-hedging in practice.
3.  Build a complete P&L attribution framework — Delta, Gamma, Vega, Vanna, and Theta — and verify it accurately explains overnight P&L under a realistic market move.
4.  Critically evaluate where and why the linear/quadratic approximation breaks down as spot moves away from current market.

# Exposure

```{r exposure_table, echo=FALSE}
dplyr::tribble(
  ~Position, ~Type,  ~`Strike K ($/bbl)`, ~Maturity, ~`Impl. Vol`, ~Contracts,
  "P1", "Put",  70.00, "3M", "35.00%", "+2,000 long",
  "P2", "Put",  55.00, "3M", "42.00%", "-2,000 short",
  "P3", "Call", 80.00, "3M", "30.00%", "-2,000 short"
) |>
  gt::gt() |>
  gt::tab_header(
    title    = "WTI Options Portfolio",
    subtitle = "Three-way collar: long put spread + short OTM call  |  F\u2080 = $73/bbl"
  ) |>
  gt::fmt_number(columns = `Strike K ($/bbl)`, decimals = 2)
```

**Market parameters**: current WTI futures price F₀ = **\$73.00/bbl**, risk-free rate r = **4.00%** (annualised, continuously compounded).

Select the appropriate pricing model from the `RTL` package for options on WTI crude oil futures. Consider the nature of the underlying contract — a NYMEX CL futures — when choosing the model and setting its cost-of-carry parameter.

# Question 1 (9 points)

For **each position**, compute the following Greeks **numerically** using central differences evaluated at `F_0`:

-   **Delta** (∂V/∂F)
-   **Gamma** (∂²V/∂F²)
-   **Vega** (∂V/∂σ)
-   **Theta** (∂V/∂T)

**Step size selection**: For each Greek, choose an appropriate finite-difference step size and explain your reasoning. A good step is small enough that the finite difference accurately approximates the true derivative, but not so small that floating-point subtraction errors dominate. Consider the natural scale of each risk factor (price in \$/bbl, vol as a decimal, time in years) and any special considerations for the short-dated options in this portfolio.

**Dollar scaling**: Each raw numerical derivative gives the option price change **per barrel** for a unit move in the risk factor. To express the Greek as a total dollar P&L impact for a position, multiply by `n × mult`, where `n` is the signed contract count (positive = long, negative = short) and `mult = 1,000 bbl/contract`. For example, a delta of 0.50/bbl for a 3,000-contract long position means the desk gains or loses 0.50 × 3,000 × 1,000 = \$1,500,000 per \$1/bbl move in WTI.

Create plots showing each Greek vs. futures price over F ∈ \[50, 100\], with one line per position.

**d.** Compute the **portfolio mark-to-market (MTM)** value and **portfolio Gamma** across F ∈ \[50, 100\]. Plot them on two separate panels. In 3–5 sentences, explain:

-   What the MTM profile shape reveals about the portfolio's net directionality and curvature.
-   Where portfolio Gamma is highest and lowest across the price range, and why those regions correspond to specific positions.
-   What the changing Gamma means for a delta-hedger rebalancing daily as WTI moves through the \$50–\$100 range.

```{r q1}
# Step sizing
delta_step <- 0.01 # Minimum change in price on CL contract (As per CME CL contract specs)
vega_step <- 0.01 # 1% change in implied vol
theta_step <- 1/365 # 1 day (in years), 

# -------------------------------------------------------------------------------------------------------
position_greeks <- function(F_0, X, T2M, r, b = 0, sigma, step_F, step_S, step_T, type){
  # Gets the greeks at whatever the current Spot price is, given by F_0
  # Gotta run ts in another function to get portfolio greeks and shit
  # F_0: Spot
  # X: strike
  # T2M: time 2 maturity
  # r: rf rate
  # b: cost of carry
  # sigma: implied vol
  # step_F: step size for the spot
  # step_S: step size for vol
  # step_T: step size in time
  # type; call or put buster
  
  # Price at current spot
  current_price <- RTL::GBSOption(F_0, X, T2M, r, b, sigma, type)$price
  
  # Delta central diff
  price_up <- RTL::GBSOption(F_0 + step_F, X, T2M, r, b, sigma, type)$price
  price_down <- RTL::GBSOption(F_0 - step_F, X, T2M, r, b, sigma, type)$price
  
  delta <- (price_up - price_down) / (2 * step_F)

  # Gamma central diffe
  gamma <- (price_up - 2 * current_price + price_down) / (step_F ^ 2)
  
  # Vega - Vol up and down
  price_vol_up <- RTL::GBSOption(F_0, X, T2M, r, b, sigma + step_S, type)$price
  price_vol_down <- RTL::GBSOption(F_0, X, T2M, r, b, sigma - step_S, type)$price
  
  vega <- (price_vol_up - price_vol_down) / (2 * step_S)
  
  # Theta - Time forward :P
  price_t_up <- RTL::GBSOption(F_0, X, T2M - step_T, r, b, sigma, type)$price
  
  theta <- (price_t_up - current_price) / step_T
  
  # Vanna
  price_up_up <- RTL::GBSOption(F_0 + step_F, X, T2M, r, b, sigma + step_S, type)$price
  price_up_dn <- RTL::GBSOption(F_0 + step_F, X, T2M, r, b, sigma - step_S, type)$price
  price_dn_up <- RTL::GBSOption(F_0 - step_F, X, T2M, r, b, sigma + step_S, type)$price
  price_dn_dn <- RTL::GBSOption(F_0 - step_F, X, T2M, r, b, sigma - step_S, type)$price
  
  vanna <- (price_up_up - price_up_dn - price_dn_up + price_dn_dn) / (4 * step_F * step_S)
  
  
  
  price <- current_price # renaming this before putting it in the list, keep it pretty
  
  # Output list (KEEP THIS IN THE SAME ORDER OR THERE WILL BE CONSEQUENCES BRUH)
  greeks <- list(delta, gamma, vega, theta, vanna, price)
  
  return(greeks)
  
}

portfolio_greeks <- function(portfolio, r, b = 0, step_F, step_S, step_T, range = c(50, 100)) {
  # Gets the greeks for all the possible combinations of underlying spot price
  # Portfolio: port provided by phil
  # r: risk free rate
  # b: cost of carry
  # step_f: step size for futures price
  # step_s: step size for volatility
  # step_t: step size for time
  # range: range of values to test 
  
  # Number of contracts for each of the positions, keeps code below clearer
  P1_n <- portfolio$n[1]
  P2_n <- portfolio$n[2]
  P3_n <- portfolio$n[3]
  
  # Spine of the df, just has the spot value in range
  input_df <- data.frame(S = seq(range[1], range[2], by = step_F))
  
  # Grabs the list of all the greeks for the positions
  greeks_lists <- input_df %>%
    rowwise() %>%
    dplyr::mutate(P1_greeks = list(mapply(position_greeks, S, portfolio$K[1], portfolio$T2M[1], r, 0, portfolio$sigma[1], step_F, step_S, step_T, portfolio$type[1])),
                  P2_greeks = list(mapply(position_greeks, S, portfolio$K[2], portfolio$T2M[2], r, 0, portfolio$sigma[2], step_F, step_S, step_T, portfolio$type[2])),
                  P3_greeks = list(mapply(position_greeks, S, portfolio$K[3], portfolio$T2M[3], r, 0, portfolio$sigma[3], step_F, step_S, step_T, portfolio$type[3])))
  
  
  # This shit gets ugly here, im sorry phil, i couldnt think of any other way to do this part lol 
  # The columns with n are weighted to be total greek for each position,
  greeks_broken <- greeks_lists %>%
    dplyr::mutate(
      # Delta B-)
      P1_delta = P1_greeks[[1]],
      P1_delta_n = P1_delta * P1_n,
      P2_delta = P2_greeks[[1]],
      P2_delta_n = P2_delta * P2_n,
      P3_delta = P3_greeks[[1]],
      P3_delta_n = P3_delta * P3_n,
      
      # Gamma !!!!!
      P1_gamma = P1_greeks[[2]],
      P1_gamma_n = P1_gamma * P1_n,
      P2_gamma = P2_greeks[[2]],
      P2_gamma_n = P2_gamma * P2_n,
      P3_gamma = P3_greeks[[2]],
      P3_gamma_n = P3_gamma * P3_n,
      
      # Vega(s) O.O
      P1_vega = P1_greeks[[3]],
      P1_vega_n = P1_vega * P1_n,
      P2_vega = P2_greeks[[3]],
      P2_vega_n = P2_vega * P2_n,
      P3_vega = P3_greeks[[3]],
      P3_vega_n = P3_vega * P3_n,
      
      # Theta 
      P1_theta = P1_greeks[[4]],
      P1_theta_n = P1_theta * P1_n,
      P2_theta = P2_greeks[[4]],
      P2_theta_n = P2_theta * P2_n,
      P3_theta = P3_greeks[[4]],
      P3_theta_n = P3_theta * P3_n,
      
      # (Ha)Vanna
      P1_vanna = P1_greeks[[5]],
      P1_vanna_n = P1_vanna * P1_n,
      P2_vanna = P2_greeks[[5]],
      P2_vanna_n = P2_vanna * P2_n,
      P3_vanna = P3_greeks[[5]],
      P3_vanna_n = P3_vanna * P3_n,
      
      # Price
      P1_price = P1_greeks[[6]],
      P1_price_n = P1_price * P1_n,
      P2_price = P2_greeks[[6]],
      P2_price_n = P2_price * P2_n,
      P3_price = P3_greeks[[6]],
      P3_price_n = P3_price * P3_n,
      
      # Portfolio crap
      Port_delta_n = P1_delta_n + P2_delta_n + P3_delta_n,
      Port_gamma_n = P1_gamma_n + P2_gamma_n + P3_gamma_n,
      Port_vega_n = P1_vega_n + P2_vega_n + P3_vega_n,
      Port_theta_n = P1_theta_n + P2_theta_n + P3_theta_n,
      Port_vanna_n = P1_vanna_n + P2_vanna_n + P3_vanna_n,
      Port_price_n = P1_price_n + P2_price_n + P3_price_n
    ) %>%
    dplyr::select(-c(P1_greeks, P2_greeks, P3_greeks))
  
  return(greeks_broken)
  
}
# -------------------------------------------------------------------------------------------------------


# The main portfolio
portfolio <- portfolio_greeks(port, r, 0, delta_step, vega_step, theta_step)



# Create plots showing each Greek vs. futures price over F ∈ [50, 100], with one line per position.
delta_plot <- plotly::plot_ly(data = portfolio) %>%
  add_trace(
    x = ~ S,
    y = ~ P1_delta_n,
    type = "scatter",
    mode = "lines",
    name = "P1"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P2_delta_n,
    type = "scatter",
    mode = "lines",
    name = "P2"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P3_delta_n,
    type = "scatter",
    mode = "lines",
    name = "P3"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ Port_delta_n,
    type = "scatter",
    mode = "lines",
    name = "Portfolio"
  ) %>%
  layout(
    title = list(text = "Delta vs Futures Price."),
    xaxis = list(title = "Futures Price"),
    yaxis = list(title = "Delta, weighted")
  )

gamma_plot <- plotly::plot_ly(data = portfolio) %>%
  add_trace(
    x = ~ S,
    y = ~ P1_gamma_n,
    type = "scatter",
    mode = "lines",
    name = "P1"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P2_gamma_n,
    type = "scatter",
    mode = "lines",
    name = "P2"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P3_gamma_n,
    type = "scatter",
    mode = "lines",
    name = "P3"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ Port_gamma_n,
    type = "scatter",
    mode = "lines",
    name = "Portfolio"
  ) %>%
  layout(
    title = list(text = "Gamma vs Futures Price."),
    xaxis = list(title = "Futures Price"),
    yaxis = list(title = "Gamma, weighted")
  )

vega_plot <- plotly::plot_ly(data = portfolio) %>%
  add_trace(
    x = ~ S,
    y = ~ P1_vega_n,
    type = "scatter",
    mode = "lines",
    name = "P1"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P2_vega_n,
    type = "scatter",
    mode = "lines",
    name = "P2"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P3_vega_n,
    type = "scatter",
    mode = "lines",
    name = "P3"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ Port_vega_n,
    type = "scatter",
    mode = "lines",
    name = "Portfolio"
  ) %>%
  layout(
    title = list(text = "Vega vs Futures Price."),
    xaxis = list(title = "Futures Price"),
    yaxis = list(title = "Vega, weighted")
  )

theta_plot <- plotly::plot_ly(data = portfolio) %>%
  add_trace(
    x = ~ S,
    y = ~ P1_theta_n,
    type = "scatter",
    mode = "lines",
    name = "P1"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P2_theta_n,
    type = "scatter",
    mode = "lines",
    name = "P2"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ P3_theta_n,
    type = "scatter",
    mode = "lines",
    name = "P3"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ Port_theta_n,
    type = "scatter",
    mode = "lines",
    name = "Portfolio"
  ) %>%
  layout(
    title = list(text = "Theta vs Futures Price."),
    xaxis = list(title = "Futures Price"),
    yaxis = list(title = "Theta, weighted")
  )

greeks_plot <- subplot(delta_plot, gamma_plot, 
                       vega_plot, theta_plot, 
                       nrows = 2, titleX = TRUE, 
                       titleY = TRUE, margin = 0.05) %>%
  layout(title = list(text = "Greeks vs Futures Price."),
         showlegend = FALSE)
# -----------------------------------------------------------------------------

# Mark to market !!!!! (part d)
pnl <- portfolio %>%
  dplyr::select(S, contains("Port")) %>%
  dplyr::mutate(PNL = Port_price_n * mult)

MTM_plot <- plotly::plot_ly(data = pnl) %>%
  add_trace(
    x = ~ S,
    y = ~ PNL,
    type = "scatter",
    mode = "lines",
    name = "Mark-to-market Value ($)"
  ) %>%
  layout(
    title = list(text = "Portfolio Mark-to-market value ($)"),
    xaxis = list(title = "Futures Price"),
    yaxis = list(title = "P&L")
  )

Port_gamma_plot <- plotly::plot_ly(data = pnl) %>%
  add_trace(
    x = ~ S,
    y = ~ Port_gamma_n * mult,
    type = "scatter",
    mode = "lines",
    name = "Portfolio Gamma"
  ) %>%
  layout(
    title = list(text = "Portfolio Gamma"),
    xaxis = list(title = "Futures Price"),
    yaxis = list(title = "Gamma P&L")
  )
# -----------------------------------------------------------------

max_gamma <- portfolio$S[portfolio$Port_gamma_n == max(portfolio$Port_gamma_n)]
min_gamma <- portfolio$S[portfolio$Port_gamma_n == min(portfolio$Port_gamma_n)]
```

# Question 2 (3 points)

Sum Greeks across positions to show **portfolio-level Greeks** at the current market. Present in a `gt::gt()` table. At approximately what futures price does the portfolio delta change sign?

```{r q2}
# Just where spot is current spot, makes shit easier for the tribble
q2 <- portfolio[portfolio$S == F_0,]

dplyr::tribble(
  ~ Position, ~ Delta, ~ Gamma, ~ Vega, ~ Theta, ~ Vanna, ~ Sum,
  "P1", q2$P1_delta_n, q2$P1_gamma_n, q2$P1_vega_n, q2$P1_theta_n, q2$P1_vanna_n, sum(q2$P1_delta_n, q2$P1_gamma_n, q2$P1_vega_n, q2$P1_theta_n, q2$P1_vanna_n),
  "P2", q2$P2_delta_n, q2$P2_gamma_n, q2$P2_vega_n, q2$P2_theta_n, q2$P2_vanna_n, sum(q2$P2_delta_n, q2$P2_gamma_n, q2$P2_vega_n, q2$P2_theta_n, q2$P2_vanna_n),
  "P3", q2$P3_delta_n, q2$P3_gamma_n, q2$P3_vega_n, q2$P3_theta_n, q2$P3_vanna_n, sum(q2$P3_delta_n, q2$P3_gamma_n, q2$P3_vega_n, q2$P3_theta_n, q2$P3_vanna_n),
  "Portfolio", q2$Port_delta_n, q2$Port_gamma_n, q2$Port_vega_n, q2$Port_theta_n, q2$Port_vanna_n, sum(q2$Port_delta_n, q2$Port_gamma_n, q2$Port_vega_n, q2$Port_theta_n, q2$Port_vanna_n)) %>%
  gt::gt() %>%
  gt::tab_header(title = "Positional, and Portfolio Greeks.")

# Delta sign change
```

# Question 3 (8 points)

Overnight, US-Iran tensions escalate sharply. The US strikes Iranian nuclear facilities; Iran retaliates by mining the Strait of Hormuz. WTI gaps up at the open:

-   WTI futures: F moves from **73.00** to **90.00** (+\$17.00/bbl)
-   Implied volatility: each position's sigma spikes by **+0.20** (20 vol points)
-   Time: **1 calendar day** passes

For each position and for the portfolio total, compute:

1.  **Actual P&L**: full revaluation at new F, new sigma, new T. P&L = (new price − old price) × n × mult.
2.  **Attributed P&L**: ΔV ≈ (Delta × ΔF) + (½ × Gamma × ΔF²) + (Vega × Δσ) + (Vanna × ΔF × Δσ) + (Theta × Δt), scaled to dollars.
3.  **Unexplained residual**: Actual − Attributed.
4.  **Attribution breakdown**: what share (%) of Attributed P&L comes from each Greek?

**Vanna** (∂²V/∂F∂σ) measures how delta changes as volatility moves. With both a large spot jump and a large vol spike coinciding, Vanna is non-negligible. Use a central-difference cross-partial:

$$\text{Vanna} = \frac{V(F+h, \sigma+k) - V(F+h, \sigma-k) - V(F-h, \sigma+k) + V(F-h, \sigma-k)}{4 \cdot h \cdot k}$$

with h = `step_S` and k = `step_v`. The Vanna P&L contribution is Vanna × ΔF × Δσ.

Present all results in a `gt::gt()` table. Which position dominates the loss? Is the Greek attribution a reliable approximation here, and why or why not?

```{r q3}
# Changes
F_1 <- 90.00
delta_F <- F_1 - F_0
delta_sigma <- 0.2
delta_t <- 1/365

# !!! For each position and for the portfolio total, compute:
# Actual P&L: full revaluation at new F, new sigma, new T. P&L = (new price − old price) × n × mult.
# Attributed P&L: ΔV ≈ (Delta × ΔF) + (½ × Gamma × ΔF²) + (Vega × Δσ) + (Vanna × ΔF × Δσ) + (Theta × Δt), scaled to dollars.
# Unexplained residual: Actual − Attributed.
# Attribution breakdown: what share (%) of Attributed P&L comes from each Greek?

# Old price shit
old <- portfolio %>%
  dplyr::filter(S == F_0) %>%
  dplyr::pull(Port_price_n) * mult

# New price (requires new table)
q3_port <- port %>%
  dplyr::mutate(
    T2M = T2M - delta_t,
    sigma = sigma + delta_sigma)

q3_portfolio <- portfolio_greeks(q3_port, r, 0, delta_step, vega_step, theta_step) %>%
  dplyr::filter(S == F_1)

new <- q3_portfolio %>%
  dplyr::pull(Port_price_n) * mult

# Actual pnl 
Actual_pnl <- new - old

# Attributed pnl
q3_attributed <- portfolio %>%
  dplyr::filter(S == F_0) %>%
  dplyr::select(contains("_n")) # To grab all the adjustd greeks 

# Finds all the moves for all positions, gets the portfolio move, then gets attributed (expected) move
# again, im sorry this is ugly but i tried to write it like 4 different ways and it didnt work :(
Attributed_pnl <- q3_attributed %>%
  dplyr::select(-contains("price")) %>%
  dplyr::mutate(
                # Delta :T
                P1_delta_move = P1_delta_n * delta_F,
                P2_delta_move = P2_delta_n * delta_F,
                P3_delta_move = P3_delta_n * delta_F,
                
                # Gamma :DD
                P1_gamma_move = 0.5 * P1_gamma_n * delta_F^2,
                P2_gamma_move = 0.5 * P2_gamma_n * delta_F^2,
                P3_gamma_move = 0.5 * P3_gamma_n * delta_F^2,
                
                # Vega :$
                P1_vega_move = P1_vega_n * delta_sigma,
                P2_vega_move = P2_vega_n * delta_sigma,
                P3_vega_move = P3_vega_n * delta_sigma,
                
                # Vanna :M
                P1_vanna_move = P1_vanna_n * delta_F * delta_sigma,
                P2_vanna_move = P2_vanna_n * delta_F * delta_sigma,
                P3_vanna_move = P3_vanna_n * delta_F * delta_sigma,
                
                # Theta :/
                P1_theta_move = P1_theta_n * delta_t,
                P2_theta_move = P2_theta_n * delta_t,
                P3_theta_move = P3_theta_n * delta_t,
                
                # Totals :>
                P1_total_move = P1_delta_move + P1_gamma_move + P1_vega_move + P1_vanna_move + P1_theta_move,
                P2_total_move = P2_delta_move + P2_gamma_move + P2_vega_move + P2_vanna_move + P2_theta_move,
                P3_total_move = P3_delta_move + P3_gamma_move + P3_vega_move + P3_vanna_move + P3_theta_move,
                across(where(is.numeric), ~ .x * mult)) %>%
  dplyr::select(contains("move")) %>%
  dplyr::mutate(delta_move = P1_delta_move + P2_delta_move + P3_delta_move,
                gamma_move = P1_gamma_move + P2_gamma_move + P3_gamma_move,
                vega_move = P1_vega_move + P2_vega_move + P3_vega_move,
                vanna_move = P1_vanna_move + P2_vanna_move + P3_vanna_move,
                theta_move = P1_theta_move + P2_theta_move + P3_theta_move,
                total_move = delta_move + gamma_move + vega_move + vanna_move + theta_move)


Unexplained_pnl <- Actual_pnl - Attributed_pnl$total_move

# Gt table time !!!! :)

a <- Attributed_pnl # to make it more readable to call
dplyr::tribble(
  ~ Position, ~ Delta, ~ Gamma, ~ Vega, ~ Vanna, ~ Theta, ~ `% Attribution`,
  "P1 num.", a$P1_delta_move, a$P1_gamma_move, a$P1_vega_move, a$P1_vanna_move, a$P1_theta_move, a$P1_total_move / a$total_move * 100,
  "P2 num.", a$P2_delta_move, a$P2_gamma_move, a$P2_vega_move, a$P2_vanna_move, a$P2_theta_move, a$P2_total_move / a$total_move * 100,
  "P3 num.", a$P3_delta_move, a$P3_gamma_move, a$P3_vega_move, a$P3_vanna_move, a$P3_theta_move, a$P3_total_move / a$total_move * 100,
  "Portfolio (%)", a$delta_move / a$total_move * 100, a$gamma_move / a$total_move * 100, a$vega_move / a$total_move * 100, a$vanna_move / a$total_move * 100, a$theta_move / a$total_move * 100, 100L
) %>% gt::gt() %>%
  gt::tab_header(title = "Portfolio Attribution Properties")
```

# Question 4 (4 points)

The Iran war scenario (ΔF = +\$17) is an extreme move. How well does the Greek approximation hold as WTI moves further from the current market?

Holding vol and time constant at their initial values, compute Actual P&L (full revaluation) and Attributed P&L (Delta + Gamma only) for the portfolio across ΔF ∈ \[–20, +20\] in \$1 increments.

**a.** Plot Actual, Attributed, and the Residual on the same chart. Mark the Iran war move (ΔF = +17) with a vertical line.

**b.** At what ΔF does the absolute residual first exceed 5% of the absolute Actual P&L? Report in a `gt::gt()` table. What does this tell you about using Greeks for risk management in extreme scenarios?

```{r q4}
# New step sizing
delta_step <- 1
range_step <- 20

# Full revaluation
revaluation_port <- portfolio_greeks(port, r, 0, delta_step, delta_sigma, delta_t, range = c(F_0 - range_step, F_0 + range_step))

revaluation_port_F_0 <- revaluation_port %>%
  dplyr::filter(S == F_0)

# Question A
old_port_val <- revaluation_port_F_0$Port_price_n * mult

# Actual Pnl
Actual_pnl_new <- revaluation_port %>%
  dplyr::select(S, Port_price_n) %>%
  dplyr::mutate(Port_pnl_act = Port_price_n * mult) %>%
  dplyr::select(S, Port_pnl_act) %>%
  dplyr::mutate(Port_pnl_act = Port_pnl_act - old_port_val)

# Attributed Pnl
delta_old <- revaluation_port_F_0$Port_delta_n
gamma_old <- revaluation_port_F_0$Port_gamma_n

Attributed_pnl_new <- data.frame(S = revaluation_port$S,
                                 delta = delta_old,
                                 gamma = gamma_old) %>%
  dplyr::mutate(delta_pnl = mult * delta * (S - F_0),
                gamma_pnl = mult * gamma * 0.5 * (S - F_0)^2,
                Port_pnl_att = delta_pnl + gamma_pnl) %>%
  dplyr::select(S, Port_pnl_att)

# Pnl crazytime df !!!!!
q4_pnl <- Actual_pnl_new %>%
  dplyr::left_join(., Attributed_pnl_new, join_by(S)) %>%
  dplyr::mutate(Unexplained_pnl = Port_pnl_act - Port_pnl_att)

# Plotly vertical line function
# Copied from https://plotly.com/r/horizontal-vertical-shapes/

vline <- function(x = 0, color = "green") {
  list(
    type = "line",
    y0 = 0,
    y1 = 1,
    yref = "paper",
    x0 = x,
    x1 = x,
    line = list(color = color, dash="line")
  )
}

# -----------------------------------------------------------

# Plot of this shiesse
q4_plot <- plotly::plot_ly(data = q4_pnl) %>%
  add_trace(
    x = ~ S,
    y = ~ Port_pnl_act,
    type = "scatter",
    mode = "lines",
    name = "Actual"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ Port_pnl_att,
    type = "scatter",
    mode = "lines",
    name = "Attributed"
  ) %>%
  add_trace(
    x = ~ S,
    y = ~ Unexplained_pnl,
    type = "scatter",
    mode = "lines",
    name = "Unexplained"
  ) %>%
  layout(
    title = list(text = "Pnl, Actual, Attributed, Unexplained"),
    xaxis = list(title = "Futures Price"),
    yaxis = list(title = "Pnl ($)"),
    shapes = list(vline(F_1, "red"))
  )
```

# Question 5 (6 points)

Discuss the key learnings and insights from a Greek risk management perspective. Your explanation must:

-   Interpret the **full revaluation MTM profile** from Q1(d): describe the portfolio's curvature structure across the \$50–\$100 range. Identify the regions of maximum and minimum Gamma. Explain what the changing Gamma means for a desk running this portfolio — specifically, how frequently delta rebalancing is needed as WTI moves through the strikes.
-   Interpret the **full revaluation P&L and Gamma profiles** from Q1(d): describe the portfolio's curvature structure and identify which positions drive positive vs negative Gamma regions.
-   Identify which position and which Greek drove the largest loss in the Iran war scenario, and explain why that was predictable from the portfolio structure.
-   Explain the role of **Vanna**: why is it non-trivial when both a large spot jump and a large vol spike occur simultaneously?
-   Explain why the Greek approximation (Q4) breaks down at extreme moves, and what this means for a real desk managing this book.

Pay particular attention to why textbook delta-hedging assumptions break down in geopolitical gap scenarios, and what risk tools a real energy desk would use instead.

*Your answer here.*
